---
title: "Characterization of the exposure to carcinogenic chemicals in Luxembourg children"
author: "Franco Catuogno"
format: html
editor: visual
---

```{r setup}
#| include: false
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

# Required packages
```{r}
library(readxl)
library(dplyr)
library(purrr)
library(stringr)
library(ggplot2)
library(tibble)
library(tidyr)
library(forcats)
library(gghalves)
library(patchwork)
library(gt)
library(dunn.test)
```

# 0. Importing datasets
```{r}
pollutants_concentrations <- read_excel("data/LuxChild256.xlsx", 
    sheet = "adapted_format") |> 
  select(!3) |> 
  mutate(across(where(is.character) & !c(1:2), ~ as.numeric(na_if(.x, "nd"))))

cancersite_carcichem_biomarkers_76 <- read_excel("exported/cancersite_carcichem_biomarkers_76.xlsx") |> 
  select(!1)

chems_molecular_weight <- read_excel("data/Chemicals_MW.xlsx", 
    sheet = "chems_MW") |> 
  select(Chemical, MW)

cross_referenced_truefalse <- read_excel("exported/cross_referenced_truefalse.xlsx")

Crossreferenced_carcinogenicity_cancersite_metabolites  <- read_excel("exported/Crossreferenced_carcinogenicity_cancersite_metabolites.xlsx")

demographics <- read_excel("data/LuxChild_Demog.xlsx") |>
  rename(AntiParasites ="Anti-parasites",
         part_id = "SampleID") |> 
  mutate(AgeGroup = case_when(
    Age < 5 ~ "Preschool children",
    Age < 10 ~ "School-age children",
    Age <=13 ~ "Preadolescents"))
```

# 1. Biomarkers of exposure to carcinogenic chemicals detected in children living in Luxembourg

## 1.1. Counting the number of biomarkers by detection frequency
```{r}
carcichem_biomarkers_lux <- inner_join(cancersite_carcichem_biomarkers_76, 
           pollutants_concentrations |> 
             mutate(detection_frequency = rowMeans(!is.na(across(3:258))) * 100) |> 
             select(!Chemical), 
           by = "CAS_RN") |> 
  relocate(detection_frequency, .after = CAS_RN) |> 
  filter(!detection_frequency == 0)

print(paste("There are",
            nrow(carcichem_biomarkers_lux),
            "biomarkers detected in at least one child,",
            nrow(carcichem_biomarkers_lux |> filter(!detection_frequency < 50)),
            "biomarkers detected in at least half of the population, and",
            nrow(carcichem_biomarkers_lux |> filter(!detection_frequency < 95)),
            "biomarkers detected in more than 95% of the population"))


```

## 1.2. Plotting detection frequencies for the list of biomarkers detected

### 1.2.1. Showing agencies
```{r}

data_fig3 <- inner_join(pollutants_concentrations|> 
               mutate(detection_frequency = rowMeans(!is.na(across(3:258))) * 100) |> 
               select(Chemical, detection_frequency) |> 
               filter(!detection_frequency == 0),
           inner_join(Crossreferenced_carcinogenicity_cancersite_metabolites |> 
                          select(Chemical, Cross_referenced_Carcinogenicity),
                      cross_referenced_truefalse,
                      by = "Chemical"),
           by = "Chemical") |> 
    filter(!Cross_referenced_Carcinogenicity == "Metabolite of several chemicals (including carcinogens)" & 
               !is.na(Cross_referenced_Carcinogenicity)) |> 
    arrange(detection_frequency) |> 
    mutate(
        # First create a list of agencies with positive classifications per row
        Agency_List = pmap_chr(
            list(IARC, ECHA, USEPA, DFG, HCIS, PubChem),
            ~ {
                agencies <- c("IARC", "ECHA", "USEPA", "DFG", "HCIS", "PubChem")
                classifications <- c(...)
                positive_agencies <- agencies[classifications %in% c("TRUE")]
                
                # Format the output based on number of agencies
                if (length(positive_agencies) == 0) {
                    return("Related to a carcinogen")
                } else if (length(positive_agencies) == 1) {
                    return(positive_agencies)
                } else if (length(positive_agencies) == 2) {
                    return(str_c(positive_agencies, collapse = " & "))
                } else {
                    return(str_c(
                        str_c(positive_agencies[-length(positive_agencies)], collapse = ", "),
                        " & ",
                        positive_agencies[length(positive_agencies)]
                    ))
                }
            }
        ),
        
        # Then create the Entity column with your specific formatting
        Entity = case_when(
            Agency_List == "Related to a carcinogen" ~ "Related to a carcinogen",
            str_count(Agency_List, "&") == 0 ~ Agency_List,  # Single agency
            str_count(Agency_List, "&") == 1 ~ Agency_List,  # Two agencies
            TRUE ~ Agency_List  # Three or more agencies
        )
    ) |> 
    select(!Agency_List)



data_fig3|> 
    ggplot(aes(x = reorder(Chemical, detection_frequency), y = detection_frequency, fill = Cross_referenced_Carcinogenicity)) +  
    geom_bar(stat = "identity") +  # Create bar plot
    scale_fill_manual(values = c("Carcinogen in humans" = "#66c2a3",
                                 "Carcinogen in other models" = "#fa8d60",
                                 "Isomer in a carcinogenic mixture" = "#e78cc4",
                                 "Parent of a carcinogen" = "#ffda2f",
                                 "Metabolite of a carcinogen" = "#a8da53"  ,
                                 "Metabolite of several carcinogens" ="#8ca1ca")) +
    #   values = c("highlight" = "#B2182B", "default" = "#F0E442"),  # Set custom colors
    #  labels = c("Carcinogen (in other models)", "Carcinogen (in humans)")  # Custom legend labels
    #) +
    geom_hline(yintercept = 95, linetype = "dashed", color = "darkgrey") +  # Horizontal line at y = 90
    geom_hline(yintercept = 50, linetype = "dashed", color = "darkgrey") +  # Horizontal line at y = 50 
    # annotate("text", x = 1, y = 98, label = "DF = 90%", color = "black", size = 3, hjust = -0.1) +  # Label for the line
    labs(x = element_blank(),
         y = "Detection Frequencies %",
         fill = "Category") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5)) +  # Center the title
    theme(axis.text.x = element_text(size = 12, angle = 0, hjust = 0.5, vjust = 0.5),  # Make x-axis labels smaller and rotate
          axis.text.y = element_text(size = 14, vjust = 0.2),
          axis.title.x = element_text(size = 14),
          axis.title.y = element_text(size = 14),
          legend.text = element_text(size = 12),
          legend.position = c(0.8, 0.2),
          legend.box.background = element_rect(color="white", size=2)
    ) +
    geom_text(aes(label = Entity), 
              position = position_nudge(y = 0.3),  # Place labels at the top of the plot
              hjust = 0,
              angle = 0, 
              size = 4.0, 
              color = "black") +  # Smaller font size and rotate vertically
    scale_y_continuous(
        breaks = seq(0, 100, by = 10),  # Y-axis ticks at intervals of 5
        limits = c(0, 140)  # Set a custom y-axis maximum value (adjust if needed)
    ) + coord_flip()


```

### 1.2.2. Showing cancer sites
```{r}

data_fig3b <- inner_join(pollutants_concentrations|> 
               mutate(detection_frequency = rowMeans(!is.na(across(3:258))) * 100) |> 
               select(Chemical, detection_frequency) |> 
               filter(!detection_frequency == 0),
           inner_join(Crossreferenced_carcinogenicity_cancersite_metabolites |> 
                          select(Chemical, Cross_referenced_Carcinogenicity),
                      cancersite_carcichem_biomarkers_76 |> 
                        select(1,4:21),
                      by = "Chemical"),
           by = "Chemical") |> 
    filter(!Cross_referenced_Carcinogenicity == "Metabolite of several chemicals (including carcinogens)" & 
               !is.na(Cross_referenced_Carcinogenicity)) |> 
    arrange(detection_frequency) |> 
    mutate(
        # First create a list of agencies with positive classifications per row
        cancersite_List = pmap_chr(
            list(Breast, Prostate, Lung, Melanoma, Bladder, NHL, Kidney, Liver, Thyroid, Stomach, Thimus, Ovary, Testis, `Multiple-myeloma`, Uterus, Nasopharynx, `Soft tissue`, Adrenal),
            ~ {
                cancersite <- c(cancersite_carcichem_biomarkers_76 |>  select(!1:3)|> colnames())
                classifications <- c(...)
                positive_cancersites <- cancersite[classifications %in% c("YES")]
                
                # Format the output based on number of agencies
                if (length(positive_cancersites) == 0) {
                    return("Related to carcinogens with different cancer sites")
                } else if (length(positive_cancersites) == 1) {
                    return(positive_cancersites)
                } else if (length(positive_cancersites) == 2) {
                    return(str_c(positive_cancersites, collapse = " & "))
                } else {
                    return(str_c(
                        str_c(positive_cancersites[-length(positive_cancersites)], collapse = ", "),
                        " & ",
                        positive_cancersites[length(positive_cancersites)]
                    ))
                }
            }
        ),
        
        # Then create the Entity column with your specific formatting
        Cancer_sites = case_when(
            cancersite_List == NA ~ NA,
            str_count(cancersite_List, "&") == 0 ~ cancersite_List,  # Single agency
            str_count(cancersite_List, "&") == 1 ~ cancersite_List,  # Two agencies
            TRUE ~ cancersite_List  # Three or more agencies
        )
    ) |> 
    select(!cancersite_List)


data_fig3b|> 
    ggplot(aes(x = reorder(Chemical, detection_frequency), y = detection_frequency, fill = Cross_referenced_Carcinogenicity)) +  
    geom_bar(stat = "identity") +  # Create bar plot
    scale_fill_manual(values = c("Carcinogen in humans" = "#66c2a3",
                                 "Carcinogen in other models" = "#fa8d60",
                                 "Isomer in a carcinogenic mixture" = "#8ca1ca",
                                 "Parent of a carcinogen" = "#ffda2f",
                                 "Metabolite of a carcinogen" = "#a8da53"  ,
                                 "Metabolite of several carcinogens" = "#e78cc4")) +
    geom_hline(yintercept = 95, linetype = "dashed", color = "darkgrey") +  # Horizontal line at y = 90
    geom_hline(yintercept = 50, linetype = "dashed", color = "darkgrey") +  # Horizontal line at y = 50 
    # annotate("text", x = 1, y = 98, label = "DF = 90%", color = "black", size = 3, hjust = -0.1) +  # Label for the line
    labs(x = element_blank(),
         y = "Detection Frequencies %",
         fill = "Category") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5)) +  # Center the title
    theme(axis.text.x = element_text(size = 12, angle = 0, hjust = 0.5, vjust = 0.5),  # Make x-axis labels smaller and rotate
          axis.text.y = element_text(size = 14, vjust = 0.2),
          axis.title.x = element_text(size = 14),
          axis.title.y = element_text(size = 14),
          legend.text = element_text(size = 12),
          legend.position = c(0.8, 0.2),
          legend.box.background = element_rect(color="white", size=2)
    ) +
    geom_text(aes(label = Cancer_sites), 
              position = position_nudge(y = 0.3),  # Place labels at the top of the plot
              hjust = 0,
              angle = 0, 
              size = 4.0, 
              color = "black") +  # Smaller font size and rotate vertically
    scale_y_continuous(
        breaks = seq(0, 100, by = 10),  # Y-axis ticks at intervals of 5
        limits = c(0, 140)  # Set a custom y-axis maximum value (adjust if needed)
    ) + coord_flip()
```


# 2. Number of cancer sites where these chemicals can have an effect
```{r}
n_biomarkers_cancersite <- carcichem_biomarkers_lux |> 
  summarize(
    across(5:22,
      ~ sum(.x == "YES", na.rm = TRUE)
      )
    ) |> 
  t() |> 
  as.data.frame() |> 
  rename(n_biomarkers = "V1") |> 
  filter(!n_biomarkers == 0)

print(paste("There are",
            nrow(n_biomarkers_cancersite),
            "cancer sites where the biomarkers detected in the studied population can have an effect"))
```

# 3. Estimation of the burden by computing molar sum

## 3.1. Creating vectors of chemicals for each cancer site to be used in the molar sum
```{r}
overall_carcichem_biomarkers <- cancersite_carcichem_biomarkers_76 |>
  pull(Chemical)

breast_carcichem_biomarkers <- cancersite_carcichem_biomarkers_76 |> 
  filter(Breast == "YES") |> 
  pull(Chemical)

prostate_carcichem_biomarkers <- cancersite_carcichem_biomarkers_76 |> 
  filter(Prostate == "YES") |> 
  pull(Chemical)

lung_carcichem_biomarkers <- cancersite_carcichem_biomarkers_76 |> 
  filter(Lung == "YES") |> 
  pull(Chemical)

melanoma_carcichem_biomarkers <- cancersite_carcichem_biomarkers_76 |> 
  filter(Melanoma == "YES") |> 
  pull(Chemical)

bladder_carcichem_biomarkers <- cancersite_carcichem_biomarkers_76 |> 
  filter(Bladder == "YES") |> 
  pull(Chemical)

nhl_carcichem_biomarkers <- cancersite_carcichem_biomarkers_76 |> 
  filter(NHL == "YES") |> 
  pull(Chemical)

kidney_carcichem_biomarkers <- cancersite_carcichem_biomarkers_76 |> 
  filter(Kidney == "YES") |> 
  pull(Chemical)

liver_carcichem_biomarkers <- cancersite_carcichem_biomarkers_76 |> 
  filter(Liver == "YES") |> 
  pull(Chemical)

thyroid_carcichem_biomarkers <- cancersite_carcichem_biomarkers_76 |> 
  filter(Thyroid == "YES") |> 
  pull(Chemical)

stomach_carcichem_biomarkers <- cancersite_carcichem_biomarkers_76 |> 
  filter(Stomach == "YES") |> 
  pull(Chemical)

ovary_carcichem_biomarkers <- cancersite_carcichem_biomarkers_76 |> 
  filter(Ovary == "YES") |> 
  pull(Chemical)

testis_carcichem_biomarkers <- cancersite_carcichem_biomarkers_76 |> 
  filter(Testis == "YES") |> 
  pull(Chemical)

multiplemyeloma_carcichem_biomarkers <- cancersite_carcichem_biomarkers_76 |> 
  filter(`Multiple-myeloma` == "YES") |> 
  pull(Chemical)

uterus_carcichem_biomarkers <- cancersite_carcichem_biomarkers_76 |> 
  filter(Uterus == "YES") |> 
  pull(Chemical)

nasopharynx_carcichem_biomarkers <- cancersite_carcichem_biomarkers_76 |> 
  filter(Nasopharynx == "YES") |> 
  pull(Chemical)

soft_tissue_sarcoma_carcichem_biomarkers <- cancersite_carcichem_biomarkers_76 |> 
  filter(`Soft tissue`== "YES") |> 
  pull(Chemical)

adrenal_carcichem_biomarkers <- cancersite_carcichem_biomarkers_76 |> 
  filter(Adrenal == "YES") |> 
  pull(Chemical)

adrenal_carcichem_biomarkers <- cancersite_carcichem_biomarkers_76 |> 
  filter(Adrenal == "YES") |> 
  pull(Chemical)

thimus_carcichem_biomarkers <- cancersite_carcichem_biomarkers_76 |> 
  filter(Thimus == "YES") |> 
  pull(Chemical)

```


## 3.2. Merging concentrations with molecular weight and computing molarity and detection frequencies
```{r}
pollutants_molar_conc <- inner_join(chems_molecular_weight, pollutants_concentrations, by = "Chemical")|> 
  mutate(across(4:259)*1000/MW) |> 
  mutate(detection_frequency = rowMeans(!is.na(across(4:259))) * 100) |> 
  relocate(detection_frequency, .after = MW) |> 
  relocate(CAS_RN, .after= Chemical)

```

## 3.3. Computing molar sum including all biomarkers detected at least once and treating non-detected values as zero (Approach 3)
```{r}
pollutants_molar_conc_det <- pollutants_molar_conc |>
  filter(!detection_frequency == 0)

 pollutants_molar_conc_det_t <- pollutants_molar_conc_det|> 
  select(5:260) |> 
  t() |> 
  as.data.frame() |> 
  set_names(pollutants_molar_conc_det |> 
              pull(Chemical)) |>  
  rownames_to_column(var = "part_id") 


molarsum_carcichem_biomarkers_all <- pollutants_molar_conc_det_t|> 
  rowwise() |> 
  mutate(Σoverall_carcichem_biomarkers = sum(c_across(any_of(overall_carcichem_biomarkers)), na.rm = TRUE),
         Σbreast_carcichem_biomarkers = sum(c_across(any_of(breast_carcichem_biomarkers)), na.rm = TRUE),
         Σprostate_carcichem_biomarkers = sum(c_across(any_of(prostate_carcichem_biomarkers)), na.rm = TRUE),
         Σlung_carcichem_biomarkers = sum(c_across(any_of(lung_carcichem_biomarkers)), na.rm = TRUE),
         Σmelanoma_carcichem_biomarkers = sum(c_across(any_of(melanoma_carcichem_biomarkers)), na.rm = TRUE),
         Σbladder_carcichem_biomarkers = sum(c_across(any_of(bladder_carcichem_biomarkers)), na.rm = TRUE),
         Σnhl_carcichem_biomarkers = sum(c_across(any_of(nhl_carcichem_biomarkers)), na.rm = TRUE),
         Σkidney_carcichem_biomarkers = sum(c_across(any_of(kidney_carcichem_biomarkers)), na.rm = TRUE),
         Σliver_carcichem_biomarkers = sum(c_across(any_of(liver_carcichem_biomarkers)), na.rm = TRUE),
         Σthyroid_carcichem_biomarkers = sum(c_across(any_of(thyroid_carcichem_biomarkers)), na.rm = TRUE),
         Σstomach_carcichem_biomarkers = sum(c_across(any_of(stomach_carcichem_biomarkers)), na.rm = TRUE),
         Σovary_carcichem_biomarkers = sum(c_across(any_of(ovary_carcichem_biomarkers)), na.rm = TRUE),
         Σtestis_carcichem_biomarkers = sum(c_across(any_of(testis_carcichem_biomarkers)), na.rm = TRUE),
         Σmultiplemyeloma_carcichem_biomarkers = sum(c_across(any_of(multiplemyeloma_carcichem_biomarkers)), na.rm = TRUE),
         Σuterus_carcichem_biomarkers = sum(c_across(any_of(uterus_carcichem_biomarkers)), na.rm = TRUE),
         Σnasopharynx_carcichem_biomarkers = sum(c_across(any_of(nasopharynx_carcichem_biomarkers)), na.rm = TRUE),
         Σsoft_tissue_sarcoma_carcichem_biomarkers = sum(c_across(any_of(soft_tissue_sarcoma_carcichem_biomarkers)), na.rm = TRUE),
         Σadrenal_carcichem_biomarkers = sum(c_across(any_of(adrenal_carcichem_biomarkers)), na.rm = TRUE),
         Σthimus_carcichem_biomarkers = sum(c_across(any_of(thimus_carcichem_biomarkers)), na.rm = TRUE)
         ) |>
      select(part_id, starts_with("Σ"))


print(paste("There are",
            nrow(pollutants_molar_conc_det),
            "chemicals detected in at least one participant, among which",
            ncol(pollutants_molar_conc_det_t|> select(any_of(overall_carcichem_biomarkers))),
            "are biomarkers of exposure to carcinogens"))
```

## 3.4. Coding the boxplot
```{r}
data_fig4a <- molarsum_carcichem_biomarkers_all |> 
  select(3:20) |> 
  t() |> 
  as.data.frame() |> 
  set_names(molarsum_carcichem_biomarkers_all |> 
              pull(part_id)) |>  
  rownames_to_column(var = "cancer_site") |>
  rowwise() |> 
  mutate(median_molarsum = median(c_across(2:257))) |> 
  arrange(desc(median_molarsum))|> 
  mutate(
    cancer_site = str_remove(cancer_site, "^Σ"),
    cancer_site = str_remove(cancer_site, "_carcichem_biomarkers$"))|>
  mutate(cancer_site = str_to_sentence(cancer_site, locale = "en"),
         cancer_site = case_when(str_detect(cancer_site, regex("Multiplemyeloma", ignore_case = TRUE)) ~ "Multiple myeloma", 
                                 TRUE ~ cancer_site),
         cancer_site = case_when(str_detect(cancer_site, regex("soft_", ignore_case = TRUE)) ~ "Soft tissue",
                                 TRUE ~ cancer_site),
         cancer_site = case_when(str_detect(cancer_site, regex("nhl", ignore_case = TRUE)) ~ "NHL",
                                 TRUE ~ cancer_site)) |> 
  pivot_longer(2:257,
               names_to = "part_id",
               values_to = "molar_sum") |> 
  filter(!cancer_site == "Thimus")|>
  filter(!cancer_site == "Stomach") |> 
  mutate(cancer_site = fct_reorder(cancer_site, molar_sum, .fun = median, .desc = FALSE)) |>
  mutate(log_molar_sum = log10(molar_sum + 1))
  
# Identify which cancer_site groups have zero/near-zero spread
iqr0 <- data_fig4a |> 
  group_by(cancer_site) |> 
  summarise(iqr = IQR(molar_sum), sd = sd(molar_sum)) |> 
  filter(iqr == 0 | sd == 0) |> 
  pull(cancer_site)


fig4a <- data_fig4a  |> 
    
    ggplot(aes(x = cancer_site, y = log_molar_sum, fill = cancer_site)) +
    
    # Half boxplot (left)
    geom_half_boxplot(
        side = "l",
        width = 0.25,
        outlier.shape = NA
    ) + 
    
    # Half violin (right)
    geom_half_violin(
        data = data_fig4a |> filter(!cancer_site %in% iqr0),
        side = "r",
        alpha = 0.5,
        color = NA,
        scale = "width"
    ) +
    
    # Jittered points (centered)
    geom_jitter(
        width = 0.08,
        alpha = 0.4,
        size = 0.5
    ) +
    
    # Mean point
    stat_summary(
        fun = mean,
        geom = "point",
        shape = 23,
        size = 2.5,
        fill = "white"
    ) +
    
    scale_y_continuous(
        breaks = log10(c(0, 1, 10, 100, 1000, 10000) + 1),
        labels = c(0, 1, 10, 100, 1000, 10000)
    ) +
    labs(y = "Molar sum of biomarkers of carcinogenic chemicals",
         x = "Cancer site") +
    
    theme_minimal(base_size = 14) +
    theme(
        axis.text.x = element_text(hjust = 0.5),
        axis.title.x = element_text(hjust = 0.5),
        axis.title.y = element_text(vjust = 0.5, hjust = 0.60, size = 16),
        legend.position = "none"
    ) +
    scale_fill_manual(
        values = c(
            "#4E79A7", "#F28E2B", "#E15759", "#76B7B2",
            "#59A14F", "#EDC948", "#B07AA1", "#FF9DA7",
            "#9C755F", "#BAB0AC", "#86BCB6", "#D37295",
            "#8CD17D", "#F1CE63", "#499894", "#B6992D",
            "#79706E", "#D4A6C8"
        ),
        name = NULL
    ) +
    coord_flip()

```


# 4. Counting the number of detected biomarkers overall and by cancer site

## 4.1. Coding the computation
```{r}

numdet_carcichem_biomarkers_all <- pollutants_molar_conc_det_t|> 
  rowwise() |> 
  mutate(numdet_overall_carcichem_biomarkers = sum(
    !is.na(c_across(any_of(overall_carcichem_biomarkers)))),
         numdet_breast_carcichem_biomarkers = sum(
           !is.na(c_across(any_of(breast_carcichem_biomarkers)))),
         numdet_prostate_carcichem_biomarkers = sum(
           !is.na(c_across(any_of(prostate_carcichem_biomarkers)))),
         numdet_lung_carcichem_biomarkers = sum(
           !is.na(c_across(any_of(lung_carcichem_biomarkers)))),
         numdet_melanoma_carcichem_biomarkers = sum(
           !is.na(c_across(any_of(melanoma_carcichem_biomarkers)))),
         numdet_bladder_carcichem_biomarkers = sum(
           !is.na(c_across(any_of(bladder_carcichem_biomarkers)))),
         numdet_nhl_carcichem_biomarkers = sum(
           !is.na(c_across(any_of(nhl_carcichem_biomarkers)))),
         numdet_kidney_carcichem_biomarkers = sum(
           !is.na(c_across(any_of(kidney_carcichem_biomarkers)))),
         numdet_liver_carcichem_biomarkers = sum(
           !is.na(c_across(any_of(liver_carcichem_biomarkers)))),
         numdet_thyroid_carcichem_biomarkers = sum(
           !is.na(c_across(any_of(thyroid_carcichem_biomarkers)))),
         numdet_stomach_carcichem_biomarkers = sum(
           !is.na(c_across(any_of(stomach_carcichem_biomarkers)))),
         numdet_ovary_carcichem_biomarkers = sum(
           !is.na(c_across(any_of(ovary_carcichem_biomarkers)))),
         numdet_testis_carcichem_biomarkers = sum(
           !is.na(c_across(any_of(testis_carcichem_biomarkers)))),
         numdet_multiplemyeloma_carcichem_biomarkers = sum(     !is.na(c_across(any_of(multiplemyeloma_carcichem_biomarkers)))),
         numdet_uterus_carcichem_biomarkers = sum(
           !is.na(c_across(any_of(uterus_carcichem_biomarkers)))),
         numdet_nasopharynx_carcichem_biomarkers = sum(
           !is.na(c_across(any_of(nasopharynx_carcichem_biomarkers)))),
         numdet_soft_tissue_sarcoma_carcichem_biomarkers = sum(
           !is.na(c_across(any_of(soft_tissue_sarcoma_carcichem_biomarkers)))),
         numdet_adrenal_carcichem_biomarkers = sum(
           !is.na(c_across(any_of(adrenal_carcichem_biomarkers)))),
         numdet_thimus_carcichem_biomarkers = sum(
           !is.na(c_across(any_of(thimus_carcichem_biomarkers))))
         ) |>
      select(part_id, starts_with("numdet_"))
```

## 4.2. Coding the boxplot
```{r}
data_fig4b <- numdet_carcichem_biomarkers_all |>
  select(3:20) |> 
  t() |> 
  as.data.frame() |> 
  set_names(numdet_carcichem_biomarkers_all |> 
              pull(part_id)) |>  
  rownames_to_column(var = "cancer_site") |>
  #rowwise() |> 
  #mutate(median_numdet = median(c_across(2:257))) |> 
  #arrange(desc(median_numdet)) |> 
  pivot_longer(2:257,
               names_to = "part_id",
               values_to = "number_detected") |> 
  mutate(
    cancer_site = str_remove(cancer_site, "^numdet_"),
    cancer_site = str_remove(cancer_site, "_carcichem_biomarkers$"))  |> 
           mutate(cancer_site = str_to_sentence(cancer_site, locale = "en"),
                  cancer_site = case_when(str_detect(cancer_site, regex("nhl", ignore_case = TRUE)) ~ "NHL", 
                                       TRUE ~ cancer_site),
                  cancer_site = case_when(str_detect(cancer_site, regex("Multiple", ignore_case = TRUE)) ~ "Multiple myeloma", 
                                       TRUE ~ cancer_site),
                  cancer_site = case_when(str_detect(cancer_site, regex("soft_", ignore_case = TRUE)) ~ "Soft tissue",
                                       TRUE ~ cancer_site)) |>  
  filter(!cancer_site == "Thimus")|>
  filter(!cancer_site == "Stomach")|>
  inner_join(data_fig4a,
             by = c("cancer_site", "part_id")) |> 
  relocate(part_id)|> 
  mutate(cancer_site = fct_reorder(cancer_site, molar_sum, .fun = median, .desc = FALSE))# |>
  #mutate(log_num_det = log10(number_detected + 1))


# Identify which cancer_site groups have zero/near-zero spread
iqr0 <- data_fig4b |> 
  group_by(cancer_site) |> 
  summarise(iqr = IQR(number_detected), sd = sd(number_detected)) |> 
  filter(iqr == 0 | sd == 0) |> 
  pull(cancer_site)




fig4b <- data_fig4b |> 
 ggplot(aes(x = cancer_site, y = number_detected, fill = cancer_site)) +

   # Half boxplot (left)
  geom_half_boxplot(
    side = "l",
    width = 0.25,
    outlier.shape = NA
  ) + 
  
  # Half violin (right)
  geom_half_violin(
    data = data_fig4b |> filter(!cancer_site %in% iqr0),
    side = "r",
    alpha = 0.5,
    color = NA,
    scale = "width"
    
  ) +

  # Jittered points (centered)
  geom_jitter(
    width = 0.08,
    alpha = 0.4,
    size = 0.5
  ) +

  # Mean point
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 23,
    size = 2.5,
    fill = "white"
  ) +

 # scale_y_log10() +
  labs(y = "Number of biomarkers detected",
       x = "Cancer site") +

  theme_minimal(base_size = 14) +
  theme(
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "none"
  ) + 
  scale_fill_manual(
    values = c(
      "#4E79A7", "#F28E2B", "#E15759", "#76B7B2",
      "#59A14F", "#EDC948", "#B07AA1", "#FF9DA7",
      "#9C755F", "#BAB0AC", "#86BCB6", "#D37295",
      "#8CD17D", "#F1CE63", "#499894", "#B6992D",
      "#79706E", "#D4A6C8"
      ),
    name = NULL
    ) +
  coord_flip()

```

# 5. Comparing the 2 biomarkers:

## 5.1. Computing median values for each cancer site
```{r}
inner_join(
  molarsum_carcichem_biomarkers_all|>
    select(2:20) |> 
    t() |> 
    as.data.frame() |> 
    set_names(molarsum_carcichem_biomarkers_all |>
                pull(part_id))|>
    rownames_to_column(var = "cancer_site") |>
    rowwise() |> 
    mutate(
      median_value = median(c_across(2:257), na.rm = TRUE),
      `Molar sum` = str_glue(
        "{signif(median(c_across(2:257), na.rm = TRUE), digit = 3)} ",
        "({signif(min(c_across(2:257), na.rm = TRUE), digit = 3)} - ",
        "{round(max(c_across(2:257), na.rm = TRUE), digit = 1)})"
        ))  |> 
    select(1,258,259)|> 
    mutate(
      cancer_site = str_remove(cancer_site, "^Σ"),
      cancer_site = str_remove(cancer_site, "_carcichem_biomarkers$"))|>
    mutate(cancer_site = str_to_sentence(cancer_site, locale = "en"),
           cancer_site = case_when(str_detect(cancer_site, regex("Multiplemyeloma", ignore_case = TRUE)) ~ "Multiple myeloma", 
                                   TRUE ~ cancer_site),
           cancer_site = case_when(str_detect(cancer_site, regex("soft_", ignore_case = TRUE)) ~ "Soft tissue",
                                   TRUE ~ cancer_site),
           cancer_site = case_when(str_detect(cancer_site, regex("nhl", ignore_case = TRUE)) ~ "NHL",
                                   TRUE ~ cancer_site))
  ,
  numdet_carcichem_biomarkers_all |>
    select(2:20) |> 
    t() |> 
    as.data.frame() |>
    set_names(numdet_carcichem_biomarkers_all |> 
                pull(part_id)) |> 
    rownames_to_column(var = "cancer_site") |>
    rowwise() |> 
    mutate(
      max_value = max(c_across(2:257), na.rm = TRUE),
      `Detected number` = str_glue(
        "{median(c_across(2:257), na.rm = TRUE)} ",
        "({min(c_across(2:257), na.rm = TRUE)} - ",
        "{max(c_across(2:257), na.rm = TRUE)})"
        ),
      `Percentage of participants` = str_glue("{round(rowMeans(across(2:257) > 0) * 100, 1)}%"))|> 
    select(1,258,259,260)|> 
    mutate(
      cancer_site = str_remove(cancer_site, "^numdet_"),
      cancer_site = str_remove(cancer_site, "_carcichem_biomarkers$"))|>
    mutate(cancer_site = str_to_sentence(cancer_site, locale = "en"),
           cancer_site = case_when(str_detect(cancer_site, regex("Multiplemyeloma", ignore_case = TRUE)) ~ "Multiple myeloma", 
                                   TRUE ~ cancer_site),
           cancer_site = case_when(str_detect(cancer_site, regex("soft_", ignore_case = TRUE)) ~ "Soft tissue",
                                   TRUE ~ cancer_site),
           cancer_site = case_when(str_detect(cancer_site, regex("nhl", ignore_case = TRUE)) ~ "NHL",
                                   TRUE ~ cancer_site))
  ,
  by = "cancer_site") |>
  rename(`Cancer site`= "cancer_site") |> 
  arrange(desc(median_value), desc(max_value))|>
#  select(1,3,5,6) |> 
  gt()
```


## 5.2. Combining the 2 plots in 3 and 4
```{r}
layout <- 'AAABB'

fig4a + fig4b
```
# 6. Grouping by age or sex and comparing parameters

## 6.1. Merging data frames
```{r}
data_fig5 <- inner_join(
  demographics |>
    select(1,4,13)
  ,
  inner_join(
    molarsum_carcichem_biomarkers_all,
    numdet_carcichem_biomarkers_all,
    by = "part_id")
  ,
  by = "part_id")

```

## 6.2. Are there differences between the molar sum or the number of detected biomarkers of exposure to carcinogenic chemicals of children from different age groups?

### 6.2.1. Kruskall-Wallis comparing overall parameters by age group

#### 6.2.1.1. Molar sum of biomarkers
```{r}
dunn_result_molarsum_age <- dunn.test(
  data_fig5$Σoverall_carcichem_biomarkers,
  data_fig5$AgeGroup,
  method = "bh",
  list = TRUE
)
dunn_table_molarsum_age <- tibble(
  Comparison = dunn_result_molarsum_age$comparisons,
  Z = dunn_result_molarsum_age$Z,
  `P value` = dunn_result_molarsum_age$P,
  `Adjusted P value` = dunn_result_molarsum_age$P.adjusted
)
dunn_table_molarsum_age |>
  gt() |>
  fmt_number(
    columns = c(Z, `P value`, `Adjusted P value`),
    decimals = 3
  ) |>
  cols_label(
    Comparison = "Molar sum comparison by Age Group",
    Z = "Z",
    `P value` = "P value",
    `Adjusted P value` = "Adjusted P value"
  ) |>
  cols_align(
    align = "center",
    columns = c(Z, `P value`, `Adjusted P value`)
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.font.size = px(11),
    column_labels.background.color = "white",
    table.border.top.width = px(2),
    table.border.bottom.width = px(2),
    table_body.vlines.width = px(0),
    data_row.padding = px(5)
  )
```

#### 6.2.1.2. Detected number of biomarkers
```{r}
dunn_result_numdet_age <- dunn.test(
  data_fig5$numdet_overall_carcichem_biomarkers,
  data_fig5$AgeGroup,
  method = "bh",
  list = TRUE
)
dunn_table_numdet_age <- tibble(
  Comparison = dunn_result_numdet_age$comparisons,
  Z = dunn_result_numdet_age$Z,
  `P value` = dunn_result_numdet_age$P,
  `Adjusted P value` = dunn_table_numdet_age$P.adjusted
)

dunn_table_numdet_age |>
  gt() |>
  fmt_number(
    columns = c(Z, `P value`, `Adjusted P value`),
    decimals = 3
  ) |>
  cols_label(
    Comparison = "Detected number by Age Group",
    Z = "Z",
    `P value` = "P value",
    `Adjusted P value` = "Adjusted P value"
  ) |>
  cols_align(
    align = "center",
    columns = c(Z, `P value`, `Adjusted P value`)
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.font.size = px(11),
    column_labels.background.color = "white",
    table.border.top.width = px(2),
    table.border.bottom.width = px(2),
    table_body.vlines.width = px(0),
    data_row.padding = px(5)
  )
```


## 6.3. Are there differences between the molar sum or the number of detected biomarkers of exposure to carcinogenic chemicals of children from different sex?

### 6.3.1. Wilcoxon rank-sum test comparing overall parameters by sex
```{r}
# Molar sum
wilcox_molarsum_sex <- wilcox.test(
  Σoverall_carcichem_biomarkers ~ Sex,
  data = data_fig5,
  exact = FALSE
)

wilcox_table_molarsum_sex <- tibble(
  Comparison = "Molar sum",
  Statistic = unname(wilcox_molarsum_sex$statistic),
  `P value` = wilcox_molarsum_sex$p.value
)


# Detected number
wilcox_numdet_sex <- wilcox.test(
  numdet_overall_carcichem_biomarkers ~ Sex,
  data = data_fig5,
  exact = FALSE
)

wilcox_table_numdet_sex <- tibble(
  Comparison = "Detected number",
  Statistic = unname(wilcox_numdet_sex$statistic),
  `P value` = wilcox_numdet_sex$p.value
)

bind_rows(wilcox_table_molarsum_sex,wilcox_table_numdet_sex)|>
  rename(Parameter = "Comparison") |> 
  gt() 
```

## 6.4. Visualization of the comparison

### 6.4.1. Coding each subfigure
```{r}
fig5a <- data_fig5 |>
    ggplot(aes(x = fct_inorder(AgeGroup), y = Σoverall_carcichem_biomarkers, fill = AgeGroup)) +
    
    
    geom_half_violin(side = "l", alpha = 0.5, color = NA) +
    geom_half_boxplot(side = "r", width = 0.25, outlier.shape = NA) +
    geom_jitter(width = 0.08, alpha = 0.4, size = 1.5) +
    stat_summary(fun = mean, geom = "point", shape = 23, size = 3, fill = "red") +
    
    scale_y_log10(
        breaks = c(10, 30, 100, 300, 1000, 3000),
        labels = scales::label_number(accuracy = NULL, big.mark = ",")
    ) +
    
    coord_cartesian(clip = "off") +
    
    # Bracket: Preschool (x=1) — School-age (x=2), at y=10
    annotate("segment", x = 1, xend = 2, y = 20, yend = 20, linewidth = 0.4) +
    annotate("segment", x = 1, xend = 1, y = 20, yend = 25, linewidth = 0.4) +
    annotate("segment", x = 2, xend = 2, y = 20, yend = 25, linewidth = 0.4) +
    annotate("text", x = 1.5, y = 15, label = "q = 0.043*", size = 3.5) +
    
    stat_summary(
        fun = median,
        geom = "text",
        aes(label = paste0("Median = ", signif(after_stat(y), 3))),
        position = position_nudge(y = 0.5, x = 0.3),
        size = 3.5
    ) +
    
    labs(y = "Molar Sum (nmol/mg)") +
    
    
    theme_minimal(base_size = 14) +
    theme(
        axis.text.x     = element_blank(),
        axis.title.x    = element_blank(),
        plot.title      = element_text(face = "bold", size = 14),
        legend.position = "none"
    ) + 
    coord_cartesian(clip = "off", ylim = c(15, 3000))


fig5b <- data_fig5 |>
    ggplot(aes(x = fct_inorder(AgeGroup), y = numdet_overall_carcichem_biomarkers, fill = AgeGroup)) +
    
    geom_half_violin(side = "l", alpha = 0.5, color = NA) +
    geom_half_boxplot(side = "r", width = 0.25, outlier.shape = NA) +
    geom_jitter(width = 0.08, alpha = 0.4, size = 1.5) +
    stat_summary(fun = mean, geom = "point", shape = 23, size = 3, fill = "red") +
    
    # Bracket 1: Preschool (x=2) — School-age (x=3), at y=15
    annotate("segment", x = 1, xend = 2, y = 17, yend = 17, linewidth = 0.4) +
    annotate("segment", x = 1, xend = 1, y = 17, yend = 17.5, linewidth = 0.4) +
    annotate("segment", x = 2, xend = 2, y = 17.5, yend = 17, linewidth = 0.4) +
    annotate("text", x = 1.5, y = 16.5, label = "q = 0.007*", size = 3.5) +
    
    # Bracket 2: School-age (x=3) — Preadolescents (x=4), at y=15
    annotate("segment", x = 2, xend = 3, y = 17, yend = 17, linewidth = 0.4) +
    annotate("segment", x = 2, xend = 2, y = 16.5, yend = 17.5, linewidth = 0.4) +
    annotate("segment", x = 3, xend = 3, y = 17, yend = 17.5, linewidth = 0.4) +
    annotate("text", x = 2.5, y = 16.5, label = "q = 0.004*", size = 3.5) +
    
    labs(y = "Number of biomarkers detected") +
    
    theme_minimal(base_size = 14) +
    theme(
        axis.text.x     = element_text(hjust = 0.5),
        axis.title.x    = element_blank(),
        plot.title      = element_text(face = "bold", size = 14),
        legend.position = "none"
    )  + 
    stat_summary(
        fun = median,
        geom = "text",
        aes(label = paste0("Median = ", signif(after_stat(y), 3))),
        position = position_nudge(y = 6, x = 0.1),
        size = 3.5,
        hjust = 0
    ) +
    scale_y_continuous(
        breaks = seq(0, 50, by = 10),  # Y-axis ticks at intervals of 5
        limits = c(15, 50)  # Set a custom y-axis maximum value (adjust if needed)
    )

fig5c <- data_fig5 |>
    ggplot(aes(x = Sex, y = Σoverall_carcichem_biomarkers, fill = Sex)) +
    
    
    geom_half_violin(side = "l", alpha = 0.5, color = NA) +
    geom_half_boxplot(side = "r", width = 0.25, outlier.shape = NA) +
    geom_jitter(width = 0.08, alpha = 0.4, size = 1.5) +
    stat_summary(fun = mean, geom = "point", shape = 23, size = 3, fill = "red") +
    
    scale_y_log10(
        breaks = c(10, 30, 100, 300, 1000, 3000),
        labels = scales::label_number(accuracy = NULL, big.mark = ",")
    ) +
    
    coord_cartesian(clip = "off") +
    
    
    labs(y = "Molar Sum (nmol/mg)") +
    
    
    theme_minimal(base_size = 14) +
    theme(
        axis.text.x     = element_blank(),
        axis.title.x    = element_blank(),
        axis.title.y    = element_blank(),
        axis.text.y     = element_blank(),
        plot.title      = element_text(face = "bold", size = 14),
        legend.position = "none"
    ) + 
    stat_summary(
        fun = median,
        geom = "text",
        aes(label = paste0("Median = ", signif(after_stat(y), 3))),
        position = position_nudge(y = 0.5, x = 0.1),
        size = 3.5,
        hjust = 0
    ) + 
    coord_cartesian(clip = "off", ylim = c(15, 3000)) +
    scale_fill_brewer(palette = "Dark2", name = NULL) + 
    scale_x_discrete(labels = c(
        "F"  = "Girls",
        "M" = "Boys]"
    ))


fig5d <- data_fig5 |>
    ggplot(aes(x = Sex, y = numdet_overall_carcichem_biomarkers, fill = Sex)) +
    
    
    geom_half_violin(side = "l", alpha = 0.5, color = NA) +
    geom_half_boxplot(side = "r", width = 0.25, outlier.shape = NA) +
    geom_jitter(width = 0.08, alpha = 0.4, size = 1.5) +
    stat_summary(fun = mean, geom = "point", shape = 23, size = 3, fill = "red") +
    
    coord_cartesian(clip = "off") +
    
    
    labs(y = "Number of biomarkers detected") +
    
    
    theme_minimal(base_size = 14) +
    theme(
        axis.text.x     = element_text(hjust = 0.5),
        axis.title.x    = element_blank(),
        axis.title.y    = element_blank(),
        axis.text.y     = element_blank(),
        plot.title      = element_text(face = "bold", size = 14),
        legend.position = "none"
    ) + 
    stat_summary(
        fun = median,
        geom = "text",
        aes(label = paste0("Median = ", signif(after_stat(y), 3))),
        position = position_nudge(y = 6, x = 0.1),
        size = 3.5,
        hjust = 0
    ) + 
    scale_fill_brewer(palette = "Dark2", name = NULL) +
    scale_y_continuous(
        breaks = seq(0, 50, by = 10),  # Y-axis ticks at intervals of 5
        limits = c(15, 50)  # Set a custom y-axis maximum value (adjust if needed)
    ) + 
    scale_x_discrete(labels = c(
        "F"  = "Girls",
        "M" = "Boys"
    ))

```

### 6.4.2. Combining subfigures
```{r}
layout <- 
'AAABB
CCCDD'

fig5a + fig5c + fig5b + fig5d
```

