---
title: "Molar sum: comparison of three approaches"
author: "Franco Catuogno"
format: html
editor: visual
---

```{r setup}
#| include: false
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

# Required packages


```{r}
library(readxl)
library(dplyr)
library(tibble)
library(stringr)
library(tidyr)
library(purrr)
library(ggplot2)
```


# 0. Importing cross-referenced carcinogenicity classification and results from Luxembourg children
```{r}
Crossreferenced_carcinogenicity_cancersite_metabolites <- read_excel("exported/Crossreferenced_carcinogenicity_cancersite_metabolites.xlsx")

chems_molecular_weight <- read_excel("data/Chemicals_MW.xlsx") |> 
  rename(Chemical = "Name") |> 
  select(Chemical, `Molecular mass`) |> 
  rename(MW = "Molecular mass")

pollutants_concentrations <- read_excel("data/LuxChild256.xlsx", 
    sheet = "adapted_format") |> 
  select(!3) |> 
  mutate(across(where(is.character) & !c(1:2), ~ as.numeric(na_if(.x, "nd"))))


cancersite_carcichem_biomarkers_75 <- read_excel("exported/cancersite_carcichem_biomarkers_75.xlsx")
```


# 1. Creating vectors of chemicals for each cancer site to be used in the molar sum
```{r}
overall_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |>
  pull(Chemical)

breast_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(Breast == "YES") |> 
  pull(Chemical)

prostate_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(Prostate == "YES") |> 
  pull(Chemical)

lung_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(Lung == "YES") |> 
  pull(Chemical)

melanoma_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(Melanoma == "YES") |> 
  pull(Chemical)

bladder_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(Bladder == "YES") |> 
  pull(Chemical)

nhl_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(NHL == "YES") |> 
  pull(Chemical)

kidney_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(Kidney == "YES") |> 
  pull(Chemical)

liver_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(Liver == "YES") |> 
  pull(Chemical)

thyroid_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(Thyroid == "YES") |> 
  pull(Chemical)

stomach_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(Stomach == "YES") |> 
  pull(Chemical)

ovary_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(Ovary == "YES") |> 
  pull(Chemical)

testis_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(Testis == "YES") |> 
  pull(Chemical)

multiplemyeloma_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(`Multiple-myeloma` == "YES") |> 
  pull(Chemical)

uterus_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(Uterus == "YES") |> 
  pull(Chemical)

nasopharynx_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(Nasopharynx == "YES") |> 
  pull(Chemical)

soft_tissue_sarcoma_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(Soft_tissue_sarcoma == "YES") |> 
  pull(Chemical)

adrenal_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(Adrenal == "YES") |> 
  pull(Chemical)

adrenal_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(Adrenal == "YES") |> 
  pull(Chemical)

thimus_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(Thimus == "YES") |> 
  pull(Chemical)

```



# 2. Computing molar concentrations
## 2.1. Merging concentrations with molecular weight and computing molarity and detection frequencies
```{r}
pollutants_molar_conc <- inner_join(chems_molecular_weight, pollutants_concentrations, by = "Chemical")|> 
  mutate(across(4:259)*1000/MW) |> 
  mutate(detection_frequency = rowMeans(!is.na(across(4:259))) * 100) |> 
  relocate(detection_frequency, .after = MW) |> 
  relocate(CAS_RN, .after= Chemical)

```

## 2.2. Approach 1:Including only biomarkers with DF > 50% and substituting non-detected values by LOD/2
```{r}
pollutants_molar_conc_halflod<- pollutants_molar_conc |> 
  filter(detection_frequency >= 50) |> 
  mutate(lod = reduce(across(5:260, as.numeric), pmin, na.rm = TRUE)) |> 
  relocate(lod, .after = detection_frequency) |> 
  mutate(across(6:261, ~ifelse(is.na(.), lod/2, as.numeric(.)))) 



pollutants_molar_conc_halflod_t <- pollutants_molar_conc_halflod|>
  select(6:261) |> 
  t() |> 
  as.data.frame() |> 
  set_names(pollutants_molar_conc_halflod |> 
              pull(Chemical)) |>  
  rownames_to_column(var = "part_id")

 all26_carcichem_biomarkers_df50 <- pollutants_molar_conc_halflod_t|> 
  select(any_of(overall_carcichem_biomarkers)) |> 
   colnames()


molarsum_carcichem_biomarkers_df50 <- pollutants_molar_conc_halflod_t|> 
  rowwise() |> 
  mutate(Σoverall_carcichem_biomarkers = sum(c_across(any_of(overall_carcichem_biomarkers)), na.rm = TRUE),
         Σbreast_carcichem_biomarkers = sum(c_across(any_of(breast_carcichem_biomarkers)), na.rm = TRUE),
         Σprostate_carcichem_biomarkers = sum(c_across(any_of(prostate_carcichem_biomarkers)), na.rm = TRUE),
         Σlung_carcichem_biomarkers = sum(c_across(any_of(lung_carcichem_biomarkers)), na.rm = TRUE),
         Σmelanoma_carcichem_biomarkers = sum(c_across(any_of(melanoma_carcichem_biomarkers)), na.rm = TRUE),
         Σbladder_carcichem_biomarkers = sum(c_across(any_of(bladder_carcichem_biomarkers)), na.rm = TRUE),
         Σnhl_carcichem_biomarkers = sum(c_across(any_of(nhl_carcichem_biomarkers)), na.rm = TRUE),
         Σkidney_carcichem_biomarkers = sum(c_across(any_of(kidney_carcichem_biomarkers)), na.rm = TRUE),
         Σliver_carcichem_biomarkers = sum(c_across(any_of(liver_carcichem_biomarkers)), na.rm = TRUE),
         Σthyroid_carcichem_biomarkers = sum(c_across(any_of(thyroid_carcichem_biomarkers)), na.rm = TRUE),
         Σstomach_carcichem_biomarkers = sum(c_across(any_of(stomach_carcichem_biomarkers)), na.rm = TRUE),
         Σovary_carcichem_biomarkers = sum(c_across(any_of(ovary_carcichem_biomarkers)), na.rm = TRUE),
         Σtestis_carcichem_biomarkers = sum(c_across(any_of(testis_carcichem_biomarkers)), na.rm = TRUE),
         Σmultiplemyeloma_carcichem_biomarkers = sum(c_across(any_of(multiplemyeloma_carcichem_biomarkers)), na.rm = TRUE),
         Σuterus_carcichem_biomarkers = sum(c_across(any_of(uterus_carcichem_biomarkers)), na.rm = TRUE),
         Σnasopharynx_carcichem_biomarkers = sum(c_across(any_of(nasopharynx_carcichem_biomarkers)), na.rm = TRUE),
         Σsoft_tissue_sarcoma_carcichem_biomarkers = sum(c_across(any_of(soft_tissue_sarcoma_carcichem_biomarkers)), na.rm = TRUE),
         Σadrenal_carcichem_biomarkers = sum(c_across(any_of(adrenal_carcichem_biomarkers)), na.rm = TRUE),
         Σthimus_carcichem_biomarkers = sum(c_across(any_of(thimus_carcichem_biomarkers)), na.rm = TRUE)
         ) |> 
  select(1,52:70)


print(paste("There are",
            nrow(pollutants_molar_conc_halflod),
            "chemicals detected in at least half of the population. among which",
            ncol(pollutants_molar_conc_halflod_t|> select(any_of(overall_carcichem_biomarkers))),
            "are biomarkers of exposure to carcinogens"))
```

```{r}

```

