---
title: "Molar sum: comparison of three approaches"
author: "Franco Catuogno"
format: html
editor: visual
---

```{r setup}
#| include: false
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

# Required packages


```{r}
library(readxl)
library(dplyr)
library(tibble)
library(stringr)
library(tidyr)
library(purrr)
library(ggplot2)
library(forcats)
library(ggridges)
library(rstatix)
```


# 0. Importing cross-referenced carcinogenicity classification and results from Luxembourg children
```{r}
chems_molecular_weight <- read_excel("data/Chemicals_MW.xlsx") |> 
  rename(Chemical = "Name") |> 
  select(Chemical, `Molecular mass`) |> 
  rename(MW = "Molecular mass")

pollutants_concentrations <- read_excel("data/LuxChild256.xlsx", 
    sheet = "adapted_format") |> 
  select(!3) |> 
  mutate(across(where(is.character) & !c(1:2), ~ as.numeric(na_if(.x, "nd"))))


cancersite_carcichem_biomarkers_75 <- read_excel("exported/cancersite_carcichem_biomarkers_75.xlsx")
```


# 1. Creating vectors of chemicals for each cancer site to be used in the molar sum
```{r}
overall_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |>
  pull(Chemical)

breast_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(Breast == "YES") |> 
  pull(Chemical)

prostate_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(Prostate == "YES") |> 
  pull(Chemical)

lung_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(Lung == "YES") |> 
  pull(Chemical)

melanoma_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(Melanoma == "YES") |> 
  pull(Chemical)

bladder_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(Bladder == "YES") |> 
  pull(Chemical)

nhl_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(NHL == "YES") |> 
  pull(Chemical)

kidney_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(Kidney == "YES") |> 
  pull(Chemical)

liver_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(Liver == "YES") |> 
  pull(Chemical)

thyroid_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(Thyroid == "YES") |> 
  pull(Chemical)

stomach_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(Stomach == "YES") |> 
  pull(Chemical)

ovary_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(Ovary == "YES") |> 
  pull(Chemical)

testis_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(Testis == "YES") |> 
  pull(Chemical)

multiplemyeloma_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(`Multiple-myeloma` == "YES") |> 
  pull(Chemical)

uterus_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(Uterus == "YES") |> 
  pull(Chemical)

nasopharynx_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(Nasopharynx == "YES") |> 
  pull(Chemical)

soft_tissue_sarcoma_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(Soft_tissue_sarcoma == "YES") |> 
  pull(Chemical)

adrenal_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(Adrenal == "YES") |> 
  pull(Chemical)

adrenal_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(Adrenal == "YES") |> 
  pull(Chemical)

thimus_carcichem_biomarkers <- cancersite_carcichem_biomarkers_75 |> 
  filter(Thimus == "YES") |> 
  pull(Chemical)

```



# 2. Computing molar concentrations
## 2.1. Merging concentrations with molecular weight and computing molarity and detection frequencies
```{r}
pollutants_molar_conc <- inner_join(chems_molecular_weight, pollutants_concentrations, by = "Chemical")|> 
  mutate(across(4:259)*1000/MW) |> 
  mutate(detection_frequency = rowMeans(!is.na(across(4:259))) * 100) |> 
  relocate(detection_frequency, .after = MW) |> 
  relocate(CAS_RN, .after= Chemical)

```

## 2.2. Approach 1: Including only biomarkers with DF > 70% and doing multiple left-censored imputation to non-detected values

### 2.2.1. Data preparation: filtering chemicals and imputing nd

#### 2.2.1.1. Exclusion of chemicals with DF < 70% and computing the minimum value and using it as the lod
```{r}
pollutants_molar_conc_long_data <- pollutants_molar_conc |> 
  filter(detection_frequency >= 70) |> 
  
  mutate(
    lod = reduce(across(5:260, as.numeric), pmin, na.rm = TRUE)
  ) |> 
  
  relocate(lod, .after = detection_frequency) |> 
  
  select(!2:5) |> 
  
  pivot_longer(
    cols = 2:257,
    names_to = "part_id",
    values_to = "molarity"
  )|>
  
  group_by(Chemical) |>
  mutate(
    lod = min(molarity, na.rm = TRUE)) |>
  ungroup()
```

#### 2.2.1.2. Defining a left-censored imputation function
```{r}

fit_left_censored_lognormal <- function(x, lod) {
  
  detected <- !is.na(x)
  
  x_detected <- x[detected]
  
  n_detected <- length(x_detected)
  n_nd <- sum(!detected)
  
  # Initial estimates from detected observations
  mu_start <- mean(log(x_detected))
  sigma_start <- sd(log(x_detected))
  
  # Negative log-likelihood
  neg_loglik <- function(par) {
    
    mu <- par[1]
    sigma <- exp(par[2])
    
    # Likelihood contribution from detected observations
    loglik_detected <- sum(
      dnorm(
        log(x_detected),
        mean = mu,
        sd = sigma,
        log = TRUE
      ) - log(x_detected)
    )
    
    # Likelihood contribution from non-detects
    p_lod <- pnorm(
      (log(lod) - mu) / sigma
    )
    
    loglik_nd <- n_nd * log(p_lod)
    
    -(loglik_detected + loglik_nd)
  }
  
  # Maximum likelihood estimation
  fit <- optim(
    par = c(
      mu_start,
      log(sigma_start)
    ),
    fn = neg_loglik,
    method = "BFGS"
  )
  
  list(
    mu = fit$par[1],
    sigma = exp(fit$par[2]),
    convergence = fit$convergence,
    logLik = -fit$value
  )
}



fit_left_censored_lognormal <- function(x, lod) {
  
  detected <- !is.na(x)
  
  x_detected <- x[detected]
  
  n_detected <- length(x_detected)
  n_nd <- sum(!detected)
  
  # Initial estimates from detected observations
  mu_start <- mean(log(x_detected))
  sigma_start <- sd(log(x_detected))
  
  # Negative log-likelihood
  neg_loglik <- function(par) {
    
    mu <- par[1]
    sigma <- exp(par[2])
    
    # Likelihood contribution from detected observations
    loglik_detected <- sum(
      dnorm(
        log(x_detected),
        mean = mu,
        sd = sigma,
        log = TRUE
      ) - log(x_detected)
    )
    
    # Likelihood contribution from non-detects
    p_lod <- pnorm(
      (log(lod) - mu) / sigma
    )
    
    loglik_nd <- n_nd * log(p_lod)
    
    -(loglik_detected + loglik_nd)
  }
  
  # Maximum likelihood estimation
  fit <- optim(
    par = c(
      mu_start,
      log(sigma_start)
    ),
    fn = neg_loglik,
    method = "BFGS"
  )
  
  list(
    mu = fit$par[1],
    sigma = exp(fit$par[2]),
    convergence = fit$convergence,
    logLik = -fit$value
  )
}


impute_left_censored <- function(x, lod, seed = NULL) {
  
  if (!is.null(seed)) {
    set.seed(seed)
  }
  
  detected <- !is.na(x)
  n_nd <- sum(!detected)
  
  # If there are no non-detects, return original data
  if (n_nd == 0) {
    return(x)
  }
  
  # Fit censored log-normal distribution
  fit <- fit_left_censored_lognormal(x = x, lod = lod)
  
  mu <- fit$mu
  sigma <- fit$sigma
  
  # Cumulative probability at the LOD
  p_lod <- pnorm((log(lod) - mu) / sigma)
  
  # Generate probabilities conditional on X < LOD
  u <- runif(n_nd, min = .Machine$double.eps, max = p_lod)
  
  # Transform probabilities into concentrations
  imputed <- exp(mu + sigma * qnorm(u))
  
  # Replace non-detects
  x[!detected] <- imputed
  
  x
}

```

#### 2.2.1.3. Multiple imputation
```{r}
n_imputations <- 5

imputed_datasets <- lapply(
  1:n_imputations,
  \(i) {
    
    pollutants_molar_conc_long_data |>
      group_by(Chemical) |>
      mutate(
        molarity_imp = impute_left_censored(
          x = molarity,
          lod = lod[1],
          seed = 1000 + i
        )
      ) |>
      ungroup()
  }
)
```

#### 2.2.1.4. Converting each Multiple left-censored imputated data frame into wide formate
```{r}
pollutants_molar_conc_leftcensored_imputed <- imputed_datasets |>
  lapply(
    \(dat) {
      dat |>
        select(Chemical, part_id, molarity_imp) |>
        pivot_wider(
          names_from = part_id,
          values_from = molarity_imp
        )
    }
  )
```

#### 2.2.1.5. Showing the top 20 rows for the 5 multiple left-censored imputated data frames
```{r}
pollutants_molar_conc_leftcensored_imputed[[1]]|> head(20)
pollutants_molar_conc_leftcensored_imputed[[2]]|> head(20)
pollutants_molar_conc_leftcensored_imputed[[3]]|> head(20)
pollutants_molar_conc_leftcensored_imputed[[4]]|> head(20)
pollutants_molar_conc_leftcensored_imputed[[5]]|> head(20)
```

### 2.2.2. Diagnosis of the imputation
#### 2.2.2.1. Checking that there are no missing values left
```{r}
na_check <- pollutants_molar_conc_leftcensored_imputed |>
  lapply(\(dat) {
    dat |>
      select(-Chemical) |>
      summarise(
        n_NA = sum(is.na(pick(everything())))
      )
  })

na_check
```

#### 2.2.2.2. Checking for non-positive values
```{r}
negative_check <- pollutants_molar_conc_leftcensored_imputed |>
  lapply(\(dat) {
    dat |>
      select(-Chemical) |>
      summarise(
        n_zero = sum(pick(everything()) == 0, na.rm = TRUE),
        n_negative = sum(pick(everything()) < 0, na.rm = TRUE)
      )
  })

negative_check
```

#### 2.2.2.3. Checking that all imputed values are <lod
```{r}
original_long <- pollutants_molar_conc_long_data |>
  select(Chemical, part_id, molarity, lod)

imputed_long <- imputed_datasets |>
  lapply(\(dat) {
    
    dat |>
      select(Chemical, part_id, molarity_imp) |>
      left_join(
        original_long,
        by = c("Chemical", "part_id")
      ) |>
      mutate(
        was_imputed = is.na(molarity)
      )
  })

lod_check <- imputed_long |>
  lapply(\(dat) {
    
    dat |>
      filter(was_imputed) |>
      summarise(
        n_imputed = n(),
        n_above_lod = sum(molarity_imp >= lod),
        max_imputed_lod_ratio = max(molarity_imp / lod)
      )
  })

lod_check

```
There should not be any imputation above the LOD and the maximum ratio should be lower than 1


#### 2.2.2.3. Checking that actual concentrations were not changed
```{r}
detected_check <- imputed_long |>
  lapply(\(dat) {
    
    dat |>
      filter(!was_imputed) |>
      summarise(
        n_detected = n(),
        n_changed = sum(molarity_imp != molarity),
        max_difference = max(
          abs(molarity_imp - molarity),
          na.rm = TRUE
        )
      )
  })

detected_check
```

#### 2.2.2.4. Checking the presence of very small values
```{r}
smallest_imputed <- imputed_long |>
  lapply(\(dat) {
    
    dat |>
      filter(was_imputed) |>
      arrange(molarity_imp) |>
      slice_head(n = 10)
  })

smallest_imputed[[1]]
```

#### 2.2.2.5. Checking if the distribution of imputed values look compatible with the observed values
```{r}
distribution_check <- imputed_long |>
  lapply(\(dat) {
    
    dat |>
      group_by(Chemical) |>
      summarise(
        n_detected = sum(!was_imputed),
        n_imputed = sum(was_imputed),
        median_detected = median(
          molarity[!was_imputed],
          na.rm = TRUE
        ),
        median_imputed = if (n_imputed > 0) {
          median(molarity_imp[was_imputed], na.rm = TRUE)
        } else {
          NA_real_
        },
        max_imputed = if (n_imputed > 0) {
          max(molarity_imp[was_imputed], na.rm = TRUE)
        } else {
          NA_real_
        },
        lod = first(lod),
        max_imputed_lod_ratio = if (n_imputed > 0) {
          max_imputed / lod
        } else {
          NA_real_
        },
        .groups = "drop"
      )
  })

distribution_check[[1]]

imputation_summary <- imputed_long[[1]] |>
  group_by(Chemical) |>
  summarise(
    detection_frequency = mean(!was_imputed) * 100,
    n_imputed = sum(was_imputed),
    .groups = "drop"
  ) |>
  arrange(detection_frequency)

imputation_summary
```

#### 2.2.2.6. Automatic flagging problematic distributions
```{r}
distribution_check_all <- distribution_check |>
  imap(\(dat, i) dat |> mutate(imputation = i)) |>
  list_rbind()

distribution_check_summary <- distribution_check_all |>
  group_by(Chemical) |>
  summarise(
    n_detected = first(n_detected),
    n_imputed = first(n_imputed),
    lod = first(lod),
    median_detected = first(median_detected),
    median_imputed_mean = mean(median_imputed, na.rm = TRUE),
    max_imputed_lod_ratio_max = max(max_imputed_lod_ratio, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    # how close is the imputed median to the detected median, relative to the LOD?
    # values near 1 mean "imputed distribution nearly touches detected distribution"
    separation_ratio = median_imputed_mean / median_detected
  )

flagged_chemicals <- distribution_check_summary |>
  mutate(
    imputation_status = case_when(
      n_imputed == 0 ~ "No non-detects - nothing imputed",
      n_imputed < 10 ~ "Few imputed values - diagnostics noisy",
      TRUE ~ "OK"
    ),
    flag_above_lod = max_imputed_lod_ratio_max > 1,
    flag_poor_separation = if_else(
      n_imputed > 0,
      separation_ratio > 0.5,
      FALSE   # nothing was imputed, so separation is meaningless, not "poor"
    ),
    any_flag = flag_above_lod | flag_poor_separation
  ) |>
  arrange(desc(any_flag), desc(separation_ratio))

flagged_chemicals |> filter(any_flag)

convergence_check <- pollutants_molar_conc_long_data |>
  group_by(Chemical) |>
  summarise(
    fit = list(fit_left_censored_lognormal(x = molarity, lod = lod[1])),
    .groups = "drop"
  ) |>
  mutate(
    convergence = map_dbl(fit, "convergence"),
    mu = map_dbl(fit, "mu"),
    sigma = map_dbl(fit, "sigma")
  ) |>
  select(-fit)

flagged_chemicals <- flagged_chemicals |>
  left_join(convergence_check, by = "Chemical") |>
  mutate(
    flag_convergence = convergence != 0,
    any_flag = any_flag | flag_convergence
  )

flagged_chemicals |> filter(any_flag)
```

#### 2.2.2.7. Visual observation of the distributions:
```{r}
chemicals_with_imputation <- distribution_check_summary |>
  filter(n_imputed > 0) |>
  pull(Chemical)

ridge_data <- imputed_long[[1]] |>
  filter(Chemical %in% chemicals_with_imputation) |>
  mutate(
    status = if_else(was_imputed, "Imputed (left-censored)", "Detected")
  )

chemical_order <- distribution_check_summary |>
  filter(Chemical %in% chemicals_with_imputation) |>
  arrange(desc(n_imputed)) |>
  pull(Chemical)

ridge_data <- ridge_data |>
  mutate(Chemical = factor(Chemical, levels = rev(chemical_order)))
```

```{r}
ggplot(
    ridge_data,
    aes(x = molarity_imp, y = Chemical, fill = status, height = after_stat(count))
) +
    geom_density_ridges(
        stat = "density",
        alpha = 0.6,
        scale = 1.5,
        rel_min_height = 0.01,
        colour = "white"
    ) +
    scale_x_log10() +
    labs(
        title = "Detected vs. imputed concentrations across chemicals",
        subtitle = "Height scaled by count, not normalized density — area reflects sample size",
        x = "Molar concentration (nmol/mg, log10 scale)",
        y = NULL,
        fill = NULL
    ) +
    theme_ridges() +
    theme(legend.position = "top")
```


### 2.2.3. Computing molar sum and pooling
#### 2.2.3.1. Molar sum in each of the 5 imputed datasets
```{r}
molarsum_carcichem_imputed <- pollutants_molar_conc_leftcensored_imputed |>
  lapply(\(dat) {
    
    dat_t <- dat |>
      select(-Chemical) |>
      t() |>
      as.data.frame() |>
      set_names(dat$Chemical) |>
      rownames_to_column(var = "part_id")
    
    dat_t |>
      rowwise() |>
      mutate(
        Σoverall_carcichem_biomarkers = sum(c_across(any_of(overall_carcichem_biomarkers)), na.rm = TRUE),
        Σbreast_carcichem_biomarkers = sum(c_across(any_of(breast_carcichem_biomarkers)), na.rm = TRUE),
        Σprostate_carcichem_biomarkers = sum(c_across(any_of(prostate_carcichem_biomarkers)), na.rm = TRUE),
        Σlung_carcichem_biomarkers = sum(c_across(any_of(lung_carcichem_biomarkers)), na.rm = TRUE),
        Σmelanoma_carcichem_biomarkers = sum(c_across(any_of(melanoma_carcichem_biomarkers)), na.rm = TRUE),
        Σbladder_carcichem_biomarkers = sum(c_across(any_of(bladder_carcichem_biomarkers)), na.rm = TRUE),
        Σnhl_carcichem_biomarkers = sum(c_across(any_of(nhl_carcichem_biomarkers)), na.rm = TRUE),
        Σkidney_carcichem_biomarkers = sum(c_across(any_of(kidney_carcichem_biomarkers)), na.rm = TRUE),
        Σliver_carcichem_biomarkers = sum(c_across(any_of(liver_carcichem_biomarkers)), na.rm = TRUE),
        Σthyroid_carcichem_biomarkers = sum(c_across(any_of(thyroid_carcichem_biomarkers)), na.rm = TRUE),
        Σstomach_carcichem_biomarkers = sum(c_across(any_of(stomach_carcichem_biomarkers)), na.rm = TRUE),
        Σovary_carcichem_biomarkers = sum(c_across(any_of(ovary_carcichem_biomarkers)), na.rm = TRUE),
        Σtestis_carcichem_biomarkers = sum(c_across(any_of(testis_carcichem_biomarkers)), na.rm = TRUE),
        Σmultiplemyeloma_carcichem_biomarkers = sum(c_across(any_of(multiplemyeloma_carcichem_biomarkers)), na.rm = TRUE),
        Σuterus_carcichem_biomarkers = sum(c_across(any_of(uterus_carcichem_biomarkers)), na.rm = TRUE),
        Σnasopharynx_carcichem_biomarkers = sum(c_across(any_of(nasopharynx_carcichem_biomarkers)), na.rm = TRUE),
        Σsoft_tissue_sarcoma_carcichem_biomarkers = sum(c_across(any_of(soft_tissue_sarcoma_carcichem_biomarkers)), na.rm = TRUE),
        Σadrenal_carcichem_biomarkers = sum(c_across(any_of(adrenal_carcichem_biomarkers)), na.rm = TRUE),
        Σthimus_carcichem_biomarkers = sum(c_across(any_of(thimus_carcichem_biomarkers)), na.rm = TRUE)
      ) |>
      ungroup() |>
      select(part_id, starts_with("Σ"))
  })
```

#### 2.2.3.2. Pooling the 5 datasets with the molar sum computed with imputed data
```{r}
molarsum_carcichem_biomarkers_df70_pooled <- molarsum_carcichem_imputed |>
  imap(\(dat, i) dat |> mutate(imputation = i)) |>
  list_rbind() |>
  group_by(part_id) |>
  summarise(
    across(
      starts_with("Σ"),
      ~mean(.x)
    ),
    .groups = "drop"
  )
  
```


## 2.3. Approach 2: Including only biomarkers with DF > 50% and substituting non-detected values by LOD/2
```{r}
pollutants_molar_conc_halflod<- pollutants_molar_conc |> 
  filter(detection_frequency >= 50) |> 
  mutate(lod = reduce(across(5:260, as.numeric), pmin, na.rm = TRUE)) |> 
  relocate(lod, .after = detection_frequency) |> 
  mutate(across(6:261, ~ifelse(is.na(.), lod/2, as.numeric(.)))) 



pollutants_molar_conc_halflod_t <- pollutants_molar_conc_halflod|>
  select(6:261) |> 
  t() |> 
  as.data.frame() |> 
  set_names(pollutants_molar_conc_halflod |> 
              pull(Chemical)) |>  
  rownames_to_column(var = "part_id")

 all26_carcichem_biomarkers_df50 <- pollutants_molar_conc_halflod_t|> 
  select(any_of(overall_carcichem_biomarkers)) |> 
   colnames()


molarsum_carcichem_biomarkers_df50 <- pollutants_molar_conc_halflod_t|> 
  rowwise() |> 
  mutate(Σoverall_carcichem_biomarkers = sum(c_across(any_of(overall_carcichem_biomarkers)), na.rm = TRUE),
         Σbreast_carcichem_biomarkers = sum(c_across(any_of(breast_carcichem_biomarkers)), na.rm = TRUE),
         Σprostate_carcichem_biomarkers = sum(c_across(any_of(prostate_carcichem_biomarkers)), na.rm = TRUE),
         Σlung_carcichem_biomarkers = sum(c_across(any_of(lung_carcichem_biomarkers)), na.rm = TRUE),
         Σmelanoma_carcichem_biomarkers = sum(c_across(any_of(melanoma_carcichem_biomarkers)), na.rm = TRUE),
         Σbladder_carcichem_biomarkers = sum(c_across(any_of(bladder_carcichem_biomarkers)), na.rm = TRUE),
         Σnhl_carcichem_biomarkers = sum(c_across(any_of(nhl_carcichem_biomarkers)), na.rm = TRUE),
         Σkidney_carcichem_biomarkers = sum(c_across(any_of(kidney_carcichem_biomarkers)), na.rm = TRUE),
         Σliver_carcichem_biomarkers = sum(c_across(any_of(liver_carcichem_biomarkers)), na.rm = TRUE),
         Σthyroid_carcichem_biomarkers = sum(c_across(any_of(thyroid_carcichem_biomarkers)), na.rm = TRUE),
         Σstomach_carcichem_biomarkers = sum(c_across(any_of(stomach_carcichem_biomarkers)), na.rm = TRUE),
         Σovary_carcichem_biomarkers = sum(c_across(any_of(ovary_carcichem_biomarkers)), na.rm = TRUE),
         Σtestis_carcichem_biomarkers = sum(c_across(any_of(testis_carcichem_biomarkers)), na.rm = TRUE),
         Σmultiplemyeloma_carcichem_biomarkers = sum(c_across(any_of(multiplemyeloma_carcichem_biomarkers)), na.rm = TRUE),
         Σuterus_carcichem_biomarkers = sum(c_across(any_of(uterus_carcichem_biomarkers)), na.rm = TRUE),
         Σnasopharynx_carcichem_biomarkers = sum(c_across(any_of(nasopharynx_carcichem_biomarkers)), na.rm = TRUE),
         Σsoft_tissue_sarcoma_carcichem_biomarkers = sum(c_across(any_of(soft_tissue_sarcoma_carcichem_biomarkers)), na.rm = TRUE),
         Σadrenal_carcichem_biomarkers = sum(c_across(any_of(adrenal_carcichem_biomarkers)), na.rm = TRUE),
         Σthimus_carcichem_biomarkers = sum(c_across(any_of(thimus_carcichem_biomarkers)), na.rm = TRUE)
         ) |> 
  select(1,52:70)


print(paste("There are",
            nrow(pollutants_molar_conc_halflod),
            "chemicals detected in at least half of the population. among which",
            ncol(pollutants_molar_conc_halflod_t|> select(any_of(overall_carcichem_biomarkers))),
            "are biomarkers of exposure to carcinogens"))
```

## 2.4. Approach 3: Including all biomarkers detected at least once and treating non-detected values as zero

```{r}
pollutants_molar_conc_t <- pollutants_molar_conc |>
  select(5:260) |> 
  t() |> 
  as.data.frame() |> 
  set_names(pollutants_molar_conc |> 
              pull(Chemical)) |>  
  rownames_to_column(var = "part_id") 


molarsum_carcichem_biomarkers_all <- pollutants_molar_conc_t|> 
  rowwise() |> 
  mutate(Σoverall_carcichem_biomarkers = sum(c_across(any_of(overall_carcichem_biomarkers)), na.rm = TRUE),
         Σbreast_carcichem_biomarkers = sum(c_across(any_of(breast_carcichem_biomarkers)), na.rm = TRUE),
         Σprostate_carcichem_biomarkers = sum(c_across(any_of(prostate_carcichem_biomarkers)), na.rm = TRUE),
         Σlung_carcichem_biomarkers = sum(c_across(any_of(lung_carcichem_biomarkers)), na.rm = TRUE),
         Σmelanoma_carcichem_biomarkers = sum(c_across(any_of(melanoma_carcichem_biomarkers)), na.rm = TRUE),
         Σbladder_carcichem_biomarkers = sum(c_across(any_of(bladder_carcichem_biomarkers)), na.rm = TRUE),
         Σnhl_carcichem_biomarkers = sum(c_across(any_of(nhl_carcichem_biomarkers)), na.rm = TRUE),
         Σkidney_carcichem_biomarkers = sum(c_across(any_of(kidney_carcichem_biomarkers)), na.rm = TRUE),
         Σliver_carcichem_biomarkers = sum(c_across(any_of(liver_carcichem_biomarkers)), na.rm = TRUE),
         Σthyroid_carcichem_biomarkers = sum(c_across(any_of(thyroid_carcichem_biomarkers)), na.rm = TRUE),
         Σstomach_carcichem_biomarkers = sum(c_across(any_of(stomach_carcichem_biomarkers)), na.rm = TRUE),
         Σovary_carcichem_biomarkers = sum(c_across(any_of(ovary_carcichem_biomarkers)), na.rm = TRUE),
         Σtestis_carcichem_biomarkers = sum(c_across(any_of(testis_carcichem_biomarkers)), na.rm = TRUE),
         Σmultiplemyeloma_carcichem_biomarkers = sum(c_across(any_of(multiplemyeloma_carcichem_biomarkers)), na.rm = TRUE),
         Σuterus_carcichem_biomarkers = sum(c_across(any_of(uterus_carcichem_biomarkers)), na.rm = TRUE),
         Σnasopharynx_carcichem_biomarkers = sum(c_across(any_of(nasopharynx_carcichem_biomarkers)), na.rm = TRUE),
         Σsoft_tissue_sarcoma_carcichem_biomarkers = sum(c_across(any_of(soft_tissue_sarcoma_carcichem_biomarkers)), na.rm = TRUE),
         Σadrenal_carcichem_biomarkers = sum(c_across(any_of(adrenal_carcichem_biomarkers)), na.rm = TRUE),
         Σthimus_carcichem_biomarkers = sum(c_across(any_of(thimus_carcichem_biomarkers)), na.rm = TRUE)
         ) |>
      select(part_id, starts_with("Σ"))

```


# 3. Comparison of the 3 approaches

## 3.1. Statistical comparison of the molar sum computed by each approach for each cancer site 

### 3.1.1. Preparing data

#### 3.1.1.1. Building the wide comparison table (unchanged)
```{r}
all3approaches_data <- molarsum_carcichem_biomarkers_df70_pooled |> 
  rename_with(~paste0(.x, "_df70_ndimp"), -part_id) |> 
  left_join(molarsum_carcichem_biomarkers_df50 |> 
              rename_with(~paste0(.x, "_df50_ndlod2"), -part_id), 
            by = "part_id") |> 
  left_join(molarsum_carcichem_biomarkers_all |> 
              rename_with(~paste0(.x, "_all_nd0"), -part_id),
            by = "part_id")

vars <- names(molarsum_carcichem_biomarkers_all)[-1]
```

#### 3.1.1.2. Creating safe helpers functions: return NA rather than erroring on zero-variance chemicals
```{r}
safe_wilcox_p <- function(a, b) {
  if (all((a - b) == 0, na.rm = TRUE)) return(NA_real_)
  tryCatch(
    wilcox.test(a, b, paired = TRUE, exact = FALSE)$p.value,
    error = function(e) NA_real_
  )
}

safe_wilcox_r <- function(a, b) {
  if (all((a - b) == 0, na.rm = TRUE)) return(NA_real_)
  tryCatch({
    eff_df <- tibble(
      val = c(a, b),
      grp = rep(c("a", "b"), each = length(a)),
      id  = rep(seq_along(a), 2)
    )
    wilcox_effsize(eff_df, val ~ grp, paired = TRUE, ref.group = "a")$effsize
  }, error = function(e) NA_real_)
}
```

#### 3.1.1.3. Coding the main loop:  Friedman + pairwise Wilcoxon
```{r}
results <- map_dfr(vars, function(v) {
  
  df <- all3approaches_data |>
    select(
      part_id,
      all_of(c(
        x1 = paste0(v, "_df70_ndimp"),
        x2 = paste0(v, "_df50_ndlod2"),
        x3 = paste0(v, "_all_nd0")
      ))
    ) |>
    filter(complete.cases(pick(x1, x2, x3)))
  
  x1 <- df$x1
  x2 <- df$x2
  x3 <- df$x3
  
  # Omnibus Friedman test
  friedman_df <- df |>
    pivot_longer(c(x1, x2, x3), names_to = "approach", values_to = "value") |>
    mutate(approach = factor(approach))
  
  friedman_p <- tryCatch(
    friedman.test(value ~ approach | part_id, data = friedman_df)$p.value,
    error = function(e) NA_real_
  )
  
  tibble(
    variable = v,
    n = nrow(df),
    p_friedman = friedman_p,
    
    # Median differences
    df70_ndimp_vs_df50_ndlod2 = median(x1 - x2, na.rm = TRUE),
    df70_ndimp_vs_all_nd0 = median(x1 - x3, na.rm = TRUE),
    df50_ndlod2_vs_all_nd0 = median(x2 - x3, na.rm = TRUE),
    
    # Wilcoxon p-values (safe)
    p_df70_ndimp_df50_ndlod2 = safe_wilcox_p(x1, x2),
    p_df70_ndimp_all_nd0 = safe_wilcox_p(x1, x3),
    p_df50_ndlod2_all_nd0 = safe_wilcox_p(x2, x3),
    
    # Effect sizes (safe)
    r_df70_ndimp_df50_ndlod2 = safe_wilcox_r(x1, x2),
    r_df70_ndimp_all_nd0 = safe_wilcox_r(x1, x3),
    r_df50_ndlod2_all_nd0 = safe_wilcox_r(x2, x3),
    
    # Direction (% higher)
    df70_ndimp_gt_df50_ndlod2 = mean(x1 > x2, na.rm = TRUE),
    df70_ndimp_gt_all_nd0 = mean(x1 > x3, na.rm = TRUE),
    df50_ndlod2_gt_all_nd0 = mean(x2 > x3, na.rm = TRUE)
  )
})
```

#### 3.1.1.4. BH correction pooled across all pairwise Wilcoxon p-values
(p_friedman is excluded from this pivot, so it survives untouched as its own column)

```{r}
results <- results |>
  pivot_longer(
    cols = c(p_df70_ndimp_df50_ndlod2, p_df70_ndimp_all_nd0, p_df50_ndlod2_all_nd0),
    names_to = "comparison",
    values_to = "p_raw"
  ) |>
  mutate(p_adj_BH = {
    p_adj <- rep(NA_real_, length(p_raw))
    non_na <- !is.na(p_raw)
    p_adj[non_na] <- p.adjust(p_raw[non_na], method = "BH")
    p_adj
  }) |>
  pivot_wider(names_from = comparison, values_from = c(p_raw, p_adj_BH)) |>
  mutate(
    friedman_sig = p_friedman < 0.05,
    sig_df70_ndimp_df50_ndlod2 = p_adj_BH_p_df70_ndimp_df50_ndlod2 < 0.05,
    sig_df70_ndimp_all_nd0 = p_adj_BH_p_df70_ndimp_all_nd0 < 0.05,
    sig_df50_ndlod2_all_nd0 = p_adj_BH_p_df50_ndlod2_all_nd0 < 0.05
  )
```


## 3.2. Plotting the differences for each comparison
```{r}
plot_df <- results |> 
  
  # reshape median differences
  pivot_longer(
    cols = c(df70_ndimp_vs_df50_ndlod2, df70_ndimp_vs_all_nd0, df50_ndlod2_vs_all_nd0),
    names_to = "comparison",
    values_to = "diff"
  ) |> 
  
  # add matching p-values
  mutate(
    p = case_when(
      comparison == "df70_ndimp_vs_df50_ndlod2" ~ p_adj_BH_p_df70_ndimp_df50_ndlod2,
      comparison == "df70_ndimp_vs_all_nd0" ~ p_adj_BH_p_df70_ndimp_all_nd0,
      comparison == "df50_ndlod2_vs_all_nd0" ~ p_adj_BH_p_df50_ndlod2_all_nd0
    ),
    comparison = recode(comparison,
                        df70_ndimp_vs_df50_ndlod2 = "df70_ndimp vs df50_ndlod2",
                        df70_ndimp_vs_all_nd0 = "df70_ndimp vs all_nd0",
                        df50_ndlod2_vs_all_nd0 = "df50_ndlod2 vs all_nd0"),
    signif = ifelse(p < 0.05, "p < 0.05", "ns")
  )

plot_df <- plot_df |> 
  mutate(
    group = ifelse(variable == "Σall60carcichems", "Total", "Cancer-specific")
  )

ggplot(plot_df, aes(x = reorder(variable, diff), y = diff, fill = signif)) +
  geom_col() +
  coord_flip() +
  
  # separate panels: comparison × group
  facet_grid(group ~ comparison, scales = "free_y", space = "free_y") +
  
  scale_fill_manual(values = c("p < 0.05" = "black", "ns" = "grey70")) +
  
  labs(
    x = "",
    y = "Median difference",
    fill = "Significance",
    title = "Comparison of Molar Sum Approaches"
  ) +
  
  theme_minimal() +
  theme(
    legend.position = "top",
    strip.text = element_text(face = "bold"),
    
    # make TOTAL stand out
    strip.text.y = element_text(size = 12, face = "bold")
  )
```

## 3.3. Bar plots to show different molar sums for each cancer site obtained by each approach

### 3.3.1. Preparing data
```{r}
medians_by_approach <- all3approaches_data |>
  select(-part_id) |>
  summarise(across(everything(), median, na.rm = TRUE)) |>
  pivot_longer(everything(), names_to = "col", values_to = "median_value") |>
  mutate(
    approach = case_when(
      str_detect(col, "_df70_ndimp$") ~ "df70_ndimp",
      str_detect(col, "_df50_ndlod2$") ~ "df50_ndlod2",
      str_detect(col, "_all_nd0$") ~ "all_nd0"
    ),
    variable = str_remove(col, "_df70_ndimp$|_df50_ndlod2$|_all_nd0$") |>
      str_remove("^Σ") |> str_remove("_carcichem_biomarkers")
  ) |>
  select(-col)

site_order <- medians_by_approach |>
  group_by(variable) |>
  summarise(avg_median = mean(median_value, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(avg_median)) |>
  pull(variable)

medians_by_approach <- medians_by_approach |>
  mutate(variable = factor(variable, levels = site_order))

medians_filtered <- medians_by_approach |> 
  group_by(variable) |> 
  filter(sum(median_value) > 0) |> 
  ungroup()

medians_filtered <- medians_filtered |>
  mutate(approach = factor(approach, levels = c("df70_ndimp", "df50_ndlod2", "all_nd0")))
```

### 3.3.2. Coding the plot
```{r}

# Lock in the same ordering used in the main plot, so bracket positions match exactly
site_levels <- medians_filtered |>
    group_by(variable) |>
    summarise(mean_median = mean(median_value), .groups = "drop") |>
    arrange(mean_median) |>
    pull(variable)

medians_filtered <- medians_filtered |>
    mutate(variable = factor(variable, levels = site_levels))

# Significant sites for df50_ndlod2 vs all_nd0, with p-value carried through
sig_df50_vs_all <- plot_df |>
    filter(comparison == "df50_ndlod2 vs all_nd0") |>
    mutate(
        variable = str_remove(variable, "^Σ") |> str_remove("_carcichem_biomarkers"),
        variable = factor(variable, levels = site_levels),
        is_sig = signif == "p < 0.05"
    ) |>
    filter(is_sig) |>
    select(variable, p)

# Bracket + label geometry
bracket_data <- medians_filtered |>
    filter(approach %in% c("all_nd0", "df50_ndlod2")) |>
    group_by(variable) |>
    summarise(max_median = max(median_value, na.rm = TRUE), .groups = "drop") |>
    inner_join(sig_df50_vs_all, by = "variable") |>
    mutate(
        y_center = as.numeric(variable),
        y_low = y_center + 0,      # df50_ndlod2 offset
        y_high = y_center + 0.3,   # all_nd0 offset
        x_bracket = max_median + 9,
        p_label = paste0("p = ", formatC(p, format = "e", digits = 1))
    )

approach_labels <- c(
    "df70_ndimp" = "1: biomarkers with DF>70% and ND imputed",
    "df50_ndlod2" = "2: biomarkers with DF>50% and ND = LOD/2",
    "all_nd0" = "3: all biomarkers and ND as 0"
)


ggplot(medians_filtered |> 
           mutate(variable = str_to_sentence(variable, locale = "en"),
                  variable = case_when(str_detect(variable, regex("nhl", ignore_case = TRUE)) ~ "NHL", 
                                       TRUE ~ variable),
                  variable = case_when(str_detect(variable, regex("soft_", ignore_case = TRUE)) ~ "Soft tissue",
                                       TRUE ~ variable)),
       aes(x = median_value, y = fct_reorder(variable, median_value, .fun = mean), fill = approach)) +
    scale_fill_manual(values = c("df70_ndimp" = "#EF6905", "df50_ndlod2" = "#F1E5A1", "all_nd0" = "#486C2F"),
                      labels = approach_labels) +
    geom_col(position = position_dodge(width = 0.9)) +
    geom_text(
        aes(label = signif(median_value, digits = 3)),
        position = position_dodge(width = 0.9),
        hjust = -0.25, vjust = 0.4, size = 3.5
    ) +
    # vertical tick down to df50_ndlod2 bar
    geom_segment(
        data = bracket_data,
        aes(x = x_bracket, xend = x_bracket, y = y_low, yend = y_high),
        inherit.aes = FALSE
    ) +
    # short horizontal ticks closing the bracket ends
    geom_segment(
        data = bracket_data,
        aes(x = x_bracket, xend = x_bracket - 0.8, y = y_low, yend = y_low),
        inherit.aes = FALSE
    ) +
    geom_segment(
        data = bracket_data,
        aes(x = x_bracket, xend = x_bracket - 0.8, y = y_high, yend = y_high),
        inherit.aes = FALSE
    ) +
    geom_text(
        data = bracket_data,
        aes(x = x_bracket + 3, y = (y_low + y_high) / 2, label = paste0("*  ", p_label)),
        inherit.aes = FALSE,
        size = 3.5, hjust = 0.1
    ) +
    labs(
        #title = "Median molar sum by cancer site and approach",
        #subtitle = "Bracket shows significant difference between approaches 3 (all biomarkers with nd = 0) and 2 (biomarkers with DF >= 50% and nd as LOD/2)(BH-adjusted p)",
        x = "Median molar sum", 
        y = "Cancer site of biomarkers of exposure", 
        fill = "Approach"
    ) +
    theme_minimal() +
    expand_limits(x = max(medians_filtered$median_value) * 1.35) +
    xlim (0,180) +
    theme(axis.text.y = element_text(size = 13),
          axis.title.x = element_text(size = 14),
          axis.title.y = element_text(size = 14),
          legend.text = element_text(size = 11),
          legend.position = c(0.80, 0.20),
          legend.box.background = element_rect(color="white", size=2),
          legend.box.margin = margin(6, 6, 6, 6))

```

