---
title: "Proposed approach for the Characterization"
author: "Franco Catuogno"
format: html
editor: visual
---

```{r setup}
#| include: false
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

# Required packages

```{r}
library(readxl)
library(dplyr)
library(readr)
library(stringr)
library(pdftools)
library(tabulapdf)
library(httr2)
library(tidyverse)
library(purrr)
library(openxlsx)
library(gt)
```

# 0. Importing list of chemicals targeted in HBRU method:

```{r}
HBRU <- read_excel("data/HBRU_targeted-chemicals.xlsx") |> 
  select(!6:7)
  #read_excel("data/PubChem.xlsx") |> 
  #select(!6) |> 
  #rename(CAS_RN = "CAS N°") 

HBRU_relatedchems <- read_excel("data/HBRU_targeted-chemicals.xlsx") |> 
  select(1:2,6:7)

```

# 1. Consulting reference agencies/data bases to identify carcinogens among the targeted chemicals.

## 1.1. Loading lists of carcinogenic chemicals from each agency/data base

### 1.1.1. IARC. Downloaded from: https://monographs.iarc.who.int/list-of-classifications/.

```{r}
IARC_agents <- read_xlsx("data/Agents Classified by the IARC Monographs, Volumes 1–142.xlsx",
          skip = 1)

IARC_chemicals <- IARC_agents |>
  rename(CAS_RN = `CAS No.`) |>
  filter(
    str_detect(Agent, "^Hexachlorocyclohexanes")|
      !is.na(CAS_RN)
  )

IARC_carcichem <- IARC_chemicals |> 
  filter(Group |> 
           str_detect("1|2A|2B"))

HBRU_IARC_carcichem <- left_join(HBRU, IARC_carcichem, by = "CAS_RN") |>
  mutate(
    Group = case_when(

      is.na(Group) &
        str_detect(Chemical, regex("HCH", ignore_case = TRUE)) ~
          IARC_carcichem$Group[
            str_detect(IARC_carcichem$Agent,
                       "^Hexachlorocyclohexanes")
          ],
        
      is.na(Group) &
        str_detect(Chemical, regex("PCB", ignore_case = TRUE)) ~
          IARC_carcichem$Group[
            str_detect(IARC_carcichem$Agent,
                       "Polychlorinated biphenyls")
          ],

      is.na(Group) &
        str_detect(Chemical, regex("cis-chlordane|trans-chlordane", ignore_case = TRUE)) ~ "Isomer in racemic mixture of Chlordane (2B)"
      ,
      
      is.na(Group) &
        str_detect(Chemical, regex("oxy-chlordane", ignore_case = TRUE)) ~ "Metabolite of several carcinogens: Chlordane (2B) and Heptachlor (2B)"
      ,
      
      is.na(Group) &
        str_detect(Chemical, regex("aldrin|dieldrin", ignore_case = TRUE)) ~
          IARC_carcichem$Group[
            str_detect(IARC_carcichem$Agent,
                       "Dieldrin")
          ],

      is.na(Group) &
        str_detect(Chemical, regex("DDT", ignore_case = TRUE)) ~
          IARC_carcichem$Group[
            str_detect(IARC_carcichem$Agent,
                       "^DDT")
          ],

      is.na(Group) &
        str_detect(Chemical, regex("IMPY", ignore_case = TRUE)) ~ "Metabolite of Diazinon (2A)",
      
            
      is.na(Group) &
        str_detect(Chemical, regex("2,4-DB", ignore_case = TRUE)) ~ "Parent of 2,4-D (2B)"
      ,
      
      is.na(Group) &
        str_detect(Chemical, regex("Atrazine desethyl", ignore_case = TRUE)) ~ "Metabolite of Atrazine (2A)",      
      
      is.na(Group) &
        str_detect(Chemical, regex("Endrin", ignore_case = TRUE)) ~ "Metabolite of several chemicals: Aldrin (2A), Dieldrin (2A) and Isodrin"
      ,
      
      is.na(Group) &
        str_detect(Chemical, regex("DMP|DMTP|DMDTP", ignore_case = TRUE)) ~ "Metabolite of several chemicals: Malathion (2A), Dimethoate and other dimethyl-OPs"
      ,
      
      is.na(Group) &
        str_detect(Chemical, regex("DEP|DETP", ignore_case = TRUE)) ~ "Metabolite of several chemicals: Parathion (2B), Diazinon (2A) and Chlorpyrifos"
      ,

      is.na(Group) &
        str_detect(Chemical, regex("DEDTP", ignore_case = TRUE)) ~ "Metabolite of several chemicals: Malathion (2A), Diazinon (2A) and Chlorpyrifos"
      ,
      
      is.na(Group) &
        str_detect(Chemical, regex("PNP", ignore_case = TRUE)) ~ "Metabolite of several chemicals: Parathion (2B), Diazinon (2A) and Chlorpyrifos"
      ,
  
      TRUE ~ Group
    )
  )  |> 
  filter(!is.na(Group)) |> 
  select(1:2,7)|> 
  rename(IARC_Carcinogenicity = "Group") |> 
  filter(!Chemical |> 
  str_detect("δ|ε"))  # removing δ-HCH & ε-HCH for which there's no evidence of carcinogenicity

  

print(paste("There are", nrow(IARC_agents), "agents listed by IARC including", nrow(IARC_chemicals), "chemicals among which", nrow(IARC_carcichem), "are carcinogenic (Group 1, 2A or 2B) or related with a carcinogen (parent/metabolite/isomer in same commercial mixture) and", nrow(HBRU_IARC_carcichem), "of them are targeted in HBRU method"))

# Comment on Hexachlorocyclohexanes: δ-HCH and ε-HCH were specifically removed because IARC report on Hexachlorocyclohexanes only mentions α-HCH and β-HCH. γ-HCH has its own category 1 for its specific CAS number: 58-89-9. 


```
"There are 1129 agents listed by IARC including 913 chemicals among which 440 are carcinogenic (Group 1, 2A or 2B) and 22 of them are targeted in HBRU method"

### 1.1.2. ECHA. Downloaded from: https://chem.echa.europa.eu/obligation-lists/clhList.

```{r}

ECHA_chemicals <- read_excel("data/Harmonised_List_2026-07-21 12_20_52.xlsx") |> 
  rename(CAS_RN = "CAS number")

ECHA_H350andH351 <- ECHA_chemicals |> 
  filter(`Hazard class, category and statement code(s)` |> 
           str_detect("H350|351")) |> 
   mutate(
    ECHA_Carcinogenicity = str_extract(
      `Hazard class, category and statement code(s)`,
      "^Carc\\.\\s*(1A|1B|2),\\s*H35[01]")) 

HBRU_ECHA_carcichem <- left_join(HBRU, ECHA_H350andH351, by = "CAS_RN") |>
   mutate(
    ECHA_Carcinogenicity = case_when(

      is.na(ECHA_Carcinogenicity) &
        str_detect(Chemical, regex("cis-chlordane|trans-chlordane", ignore_case = TRUE)) ~ "Isomer in racemic mixture of Chlordane (Carc. 2, H351)"
      ,
      
      is.na(ECHA_Carcinogenicity) &
        str_detect(Chemical, regex("oxy-chlordane", ignore_case = TRUE)) ~ "Metabolite of several carcinogens: Chlordane (Carc. 2, H351) and Heptachlor (Carc. 2, H351)"
      ,
  
      is.na(ECHA_Carcinogenicity) &
        str_detect(Chemical, regex("Metolachlor", ignore_case = TRUE)) ~
          ECHA_H350andH351$ECHA_Carcinogenicity[
            str_detect(ECHA_H350andH351$`Substance name`,
                       regex("metolachlor", ignore_case = TRUE)
                       )
          ],
      
      is.na(ECHA_Carcinogenicity) &
        str_detect(Chemical, regex("dichlorophenyl", ignore_case = TRUE)) ~ "Metabolite of Diuron (Carc. 1B, H350)"
      ,
      
      is.na(ECHA_Carcinogenicity) &
        str_detect(Chemical, regex("dichloroaniline", ignore_case = TRUE)) ~ "Metabolite of several carcinogens: Diuron (Carc. 1B, H350) and Linuron (Carc. 2, H351)"
      ,
      
      is.na(ECHA_Carcinogenicity) &
        str_detect(Chemical, regex("Endrin", ignore_case = TRUE)) ~ "Metabolite of several chemicals: Aldrin (Carc. 2, H351), Dieldrin (Carc. 2, H351) and Isodrin"
      ,
              
      TRUE ~ ECHA_Carcinogenicity
    )
   ) |> 
  select(1:2, 13) |> 
  filter(!is.na(ECHA_Carcinogenicity)) |> 
  arrange(Chemical)
  

print(paste("There are", nrow(ECHA_chemicals), "chemicals listed by ECHA including", nrow(ECHA_H350andH351), "carcinogenic chemicals (Hazard statements H350 or H351) and", nrow(HBRU_ECHA_carcichem), "of them are targeted in HBRU method"))

# Comment on chlordanes and metolachlor: Only one of the isomers has been identified as carcinogenic, however the commercial product comes as a racemic mixture of the 2 (trans/cis-chlordane and R/S-metolachlor), hence, detection of one of the isomers is a biomarker of exposure to the mixture, including the carcinogenic compound.
```

"There are 4822 chemicals listed by ECHA including 1290 carcinogenic chemicals (Hazard statements H350 or H351) and 36 of them are targeted in HBRU method"

### 1.1.3. AICIS. Downloaded from https://www.industrialchemicals.gov.au/help-and-guides/list-chemicals-high-hazards-categorisation.

```{r}
HCIS_chemicals <- read_excel("data/List_of_chemicals_with_high_hazards_for_categorisation_2025.XLSX", 
    sheet = "Consolodated List")


HCIS_H350andh351 <- read_excel("data/List_of_chemicals_with_high_hazards_for_categorisation_2025.XLSX", 
    sheet = "Consolodated List") |> 
  rename(CAS_RN = "CAS RN") |> 
  filter(`Safe Work Australia HCIS` |> 
           str_detect("H350|351"))|> 
   mutate(
    HCIS_Carcinogenicity = str_extract(
      `Safe Work Australia HCIS`,
      "^Carcinogenicity \\s*(1A|1B|2) \\s*H35[01]"))

HBRU_HCIS_carcichem <- left_join(HBRU, HCIS_H350andh351, by = "CAS_RN")|>
   mutate(
    HCIS_Carcinogenicity = case_when(

      
      is.na(HCIS_Carcinogenicity) &
        str_detect(Chemical, regex("cis-chlordane|trans-chlordane", ignore_case = TRUE)) ~ "Isomer in racemic mixture of Chlordane (Carcinogenicity 2 H351)"
      ,
      
      is.na(HCIS_Carcinogenicity) &
        str_detect(Chemical, regex("oxy-chlordane", ignore_case = TRUE)) ~ "Metabolite of several carcinogens: Chlordane (Carcinogenicity 2 H351) and Heptachlor (Carcinogenicity 2 H351)"
      ,      
      
      is.na(HCIS_Carcinogenicity) &
        str_detect(Chemical, regex("dichlorophenyl", ignore_case = TRUE)) ~ "Metabolite of Diuron (Carcinogenicity 2 H351)"
      ,
      
      is.na(HCIS_Carcinogenicity) &
        str_detect(Chemical, regex("dichloroaniline", ignore_case = TRUE)) ~ "Metabolite of several carcinogens: Diuron (Carcinogenicity 2 H351) and Linuron (Carcinogenicity 2 H351)"
      ,
      
      is.na(HCIS_Carcinogenicity) &
        str_detect(Chemical, regex("Endrin", ignore_case = TRUE)) ~ "Metabolite of several chemicals: Aldrin (Carcinogenicity 2 H351), Dieldrin (Carcinogenicity 2 H351) and Isodrin"
      ,
            
      TRUE ~ HCIS_Carcinogenicity
    )
   ) |> 
  select(1:2,10,9,20) |> 
  filter(!is.na(HCIS_Carcinogenicity)) |> 
  arrange(Chemical)
  

print(paste("There are", nrow(HCIS_chemicals), "chemicals listed by HCIS including", nrow(HCIS_H350andh351), "carcinogenic chemicals (Hazard statements H350 or H351) and", nrow(HBRU_HCIS_carcichem), "of them are targeted in HBRU method"))

print(paste("HCIS lists the same chemicals as ECHA except from:", anti_join(HBRU_ECHA_carcichem, HBRU_HCIS_carcichem, by = "Chemical") |> select(Chemical)))
```

"There are 4901 chemicals listed by HCIS including 1820 carcinogenic chemicals (Hazard statements H350 or H351) and 34 of them are targeted in HBRU method"
"HCIS lists the same chemicals as ECHA except from: c("Difenoconazole", "Metolachlor")"

### 1.1.4. PubChem. https://pubchem.ncbi.nlm.nih.gov/.

```{r}
PubChem <- read_excel("data/PubChem.xlsx")

PubChem_carcichems <- PubChem  |> 
  rename(CAS_RN = "CAS N°") |> 
  rename(PubChem_Carcinogenicity = "GHS08: Health hazard (Pubchem)") |> 
  select(1:2,6)|>
   mutate(
    PubChem_Carcinogenicity = case_when(

      
      PubChem_Carcinogenicity == "NO" &
        str_detect(Chemical, regex("cis-chlordane", ignore_case = TRUE)) ~ "Isomer in racemic mixture of Chlordane with Trans-chlordane (H351)"
      ,
      
      PubChem_Carcinogenicity == "NO" &
        str_detect(Chemical, regex("oxy-chlordane", ignore_case = TRUE)) ~ "Metabolite of several carcinogens: Trans-chlordane (H351) and Heptachlor(H351)"
      ,      

      PubChem_Carcinogenicity == "NO" &
        str_detect(Chemical, regex("dichlorophenyl", ignore_case = TRUE)) ~ "Metabolite of Diuron (H351)"
      ,      
      
      PubChem_Carcinogenicity == "NO" &
        str_detect(Chemical, regex("dichloroaniline", ignore_case = TRUE)) ~ "Metabolite of several carcinogens: Diuron (H351) and Linuron (H351)"
      ,
        
      PubChem_Carcinogenicity == "NO" &
        str_detect(Chemical, regex("Endrin", ignore_case = TRUE)) ~ "Metabolite of several carcinogens: Aldrin (H351), Dieldrin (H351) and Isodrin"
      ,     
      TRUE ~ PubChem_Carcinogenicity
    )
   )  |> 
  filter(PubChem_Carcinogenicity |> 
           str_detect("H350|351"))
  

print(paste("Among the", nrow(PubChem), "chemicals targeted in HBRU method there are", nrow(PubChem_carcichems), "chemicals with evidence of carcinogenicity (Hazard statements H350 or H351) mentioned in PubChem"))
```

### 1.1.5. DFG. Taken from https://books.publisso.de/sites/default/files/documents/series/mak/lmbv/Vol2024/Iss2/Doc002/mbwl_2024_eng.pdf.

```{r}
DFG_agents <- read_excel("data/List of MAK and BAT Values 2024.xlsx") |> 
  mutate(
    CAS_RN = str_extract(Substance, "\\d{2,7}-\\d{2}-\\d"),
    Substance_clean = Substance |>
      str_remove("\\s*\\[[^\\]]*\\]") |>
      str_squish()
  )


DFG_chemicals <- DFG_agents |>  
  filter(!is.na(CAS_RN))|>
  filter(!Substance == "1,2,3,4,5,6-Hexachlorocyclohexane techn. mixture of α-HCH [319-84-6] and β-HCH [319-85-7]")

HBRU_DFG_carcichem <- left_join(HBRU, DFG_chemicals, by = "CAS_RN") |> 
  rename(DFG_Carcinogenicity = "Category") |> 
  filter(!is.na(DFG_Carcinogenicity))

print(paste("There are", nrow(DFG_agents), "carcinogenic agents listed by DFG among which", nrow(DFG_chemicals), "are chemicals among which", nrow(HBRU_DFG_carcichem), "of them are targeted in HBRU method"))

```

"There are 385 carcinogenic agents listed by DFG among which 344 are chemicals among which 6 of them are targeted in HBRU method"

### 1.1.6. USA.

#### 1.1.6.1. IRIS. Downloaded from: https://iris.epa.gov/AtoZ/?list_type=alpha

```{r}
IRIS_chemicals <- read_excel("data/WOE_Details.xlsx") |> 
  rename(CAS_RN = "CASRN")

IRIS_carcichems <- IRIS_chemicals |>
  mutate(`WOE DESCRIPTION` = str_replace(`WOE DESCRIPTION`,
                                         "Not likely",
                                         "Not-likely"),
         `WOE DESCRIPTION` = str_replace(`WOE DESCRIPTION`,
                                         "Not classifiable",
                                         "Not-classifiable"),
         `WOE DESCRIPTION` = str_replace(`WOE DESCRIPTION`,
                                         "Not applicable",
                                         "Not-applicable")) |> 
  filter(!`WOE DESCRIPTION` |> 
           str_detect("Not-|non-|Inadequate|inadequate")) |> 
  filter(!duplicated(CAS_RN))

#IRIS_carcichems |>   arrange(`WOE DESCRIPTION`) |>   distinct(`WOE DESCRIPTION`)


HBRU_IRIS_carcichems <- left_join(HBRU, IRIS_carcichems, by = "CAS_RN") |>
  mutate(
    `WOE DESCRIPTION` = case_when(

      
      is.na(`WOE DESCRIPTION`) &
        str_detect(Chemical, regex("chlordane", ignore_case = TRUE)) ~
          IRIS_carcichems$`WOE DESCRIPTION`[
            str_detect(IRIS_carcichems$`CHEMICAL NAME`,
                       "^Chlordane")
          ],
      is.na(`WOE DESCRIPTION`) &
        str_detect(Chemical, regex("PCB", ignore_case = TRUE)) ~
          IRIS_carcichems$`WOE DESCRIPTION`[
            str_detect(IRIS_carcichems$`CHEMICAL NAME`,
                       "PCBs")
          ],
      is.na(`WOE DESCRIPTION`) &
        str_detect(Chemical, regex("o,p'-DDD", ignore_case = TRUE)) ~
          IRIS_carcichems$`WOE DESCRIPTION`[
            str_detect(IRIS_carcichems$`CHEMICAL NAME`,
                       "(DDD)")
          ],
      is.na(`WOE DESCRIPTION`) &
        str_detect(Chemical, regex("o,p'-DDE", ignore_case = TRUE)) ~
          IRIS_carcichems$`WOE DESCRIPTION`[
            str_detect(IRIS_carcichems$`CHEMICAL NAME`,
                       "(DDE)")
          ],
        is.na(`WOE DESCRIPTION`) &
        str_detect(Chemical, regex("o,p'-DDT", ignore_case = TRUE)) ~
          IRIS_carcichems$`WOE DESCRIPTION`[
            str_detect(IRIS_carcichems$`CHEMICAL NAME`,
                       "(DDT)")
          ],
      TRUE ~ `WOE DESCRIPTION`
    )
  ) |> 
  select(1:2,8)|> 
  filter(!is.na(`WOE DESCRIPTION`)) 

print(paste("There are", nrow(IRIS_chemicals), "chemicals listed by IRIS including", nrow(IRIS_carcichems), "with evidence of carcinogenicity and", nrow(HBRU_IRIS_carcichems), "of them are targeted in HBRU method"))
```

"There are 307 chemicals listed by IRIS including 143 with evidence of carcinogenicity and 26 of them are targeted in HBRU method"

#### 1.1.6.2. NPIC. Downloaded from "https://npic.orst.edu/chemicals_evaluated.pdf"

#```{r}
download.file(
  "https://npic.orst.edu/chemicals_evaluated.pdf",
  destfile = "data/chemicals_evaluated.pdf",
  mode = "wb"
)

#```

```{r}
assign_lines <- function(y, tol = 3) {
  o <- order(y)
  y_sorted <- y[o]
  grp <- cumsum(c(1, diff(y_sorted) > tol))
  out <- integer(length(y))
  out[o] <- grp
  out
}

smart_paste <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA_character_)
  out <- x[1]
  for (i in seq_along(x)[-1]) {
    out <- if (str_detect(out, "-$")) paste0(out, x[i]) else paste(out, x[i])
  }
  out
}

parse_epa_page <- function(page_number, y_tol = 3){
  page <- pdf_data("data/chemicals_evaluated.pdf")[[page_number]] |>
    filter(y > 60, !str_detect(text, "Page")) |>
    # Drop standalone header-artifact tokens FIRST, before any line clustering.
    # "CODE" and "DATE" only ever appear as the wrapped remainder of "PC CODE" /
    # "REPORT DATE" - never as real data - so exact-match removal is safe.
    filter(!text %in% c("CODE", "DATE")) |>
    mutate(line_id = assign_lines(y, tol = y_tol))
  
  # --- Drop remaining whole header/banner lines ---
  header_pattern <- paste(
    "U\\.S\\. Environmental",
    "Protection Agency",
    "CHEMICAL",
    "CAS NO\\.",
    "CANCER CLASSIFICATION",
    "REPORT",
    sep = "|"
  )
  
  line_text <- page |>
    arrange(line_id, x) |>
    group_by(line_id) |>
    summarise(line = paste(text, collapse = " "), .groups = "drop")
  
  header_ids <- line_text |>
    filter(str_detect(line, header_pattern)) |>
    pull(line_id)
  
  page <- page |> filter(!line_id %in% header_ids)
  
  # --- Column split + pivot ---
  df <- page |>
    mutate(
      column = case_when(
        x < 160 ~ "Chemical",
        x >= 160 & x < 230 ~ "CAS_RN",
        x >= 230 & x < 270 ~ "PC_Code",
        x >= 270 & x < 530 ~ "Classification",
        x >= 530 ~ "Report_Date"
      )
    ) |>
    arrange(line_id, x) |>
    group_by(line_id, column) |>
    summarise(text = paste(text, collapse = " "), .groups = "drop") |>
    pivot_wider(names_from = column, values_from = text) |>
    arrange(line_id) |>
    mutate(row_id = row_number())
  
  # --- Collapse remaining wrapped-line fragments into their record ---

df <- df |>
  mutate(
    has_pc = !is.na(PC_Code),
    grp    = rev(cumsum(rev(has_pc)))
  )

df |>
  group_by(grp) |>
  summarise(
    Chemical       = smart_paste(Chemical),
    CAS_RN         = smart_paste(CAS_RN),      # now joins "64742-89-8;" + "64742-55-8"
    PC_Code        = first(na.omit(PC_Code)),
    Classification = smart_paste(Classification),
    Report_Date    = first(na.omit(Report_Date)),
    first_row      = min(row_id),
    .groups = "drop"
  ) |>
  arrange(first_row) |>
  select(-grp, -first_row)
}

n_pages <- length(pdf_data("data/chemicals_evaluated.pdf"))

```

```{r}
USEPA_npic <- map_dfr(
  13:n_pages,
  ~ parse_epa_page(.x) |> mutate(page = .x)
  ) |> 
  filter(!is.na(Chemical))

USEPA_npic |> 
  arrange(Classification) |> 
  distinct(Classification) |> 
  filter(!str_detect(Classification, regex("^Likely To Be Carcinogenic To Humans")))

USEPA_npic_carcichems <- USEPA_npic|> 
  
  # next might be deleted later
  mutate(Classification = case_when(
    Classification == "See: Imazalil Sulfate, (PC Code 111902)." &
        str_detect(Chemical, regex("^Imazalil$", ignore_case = TRUE)) ~ "See: Imazalil Sulfate (Likely To Be Carcinogenic To Humans.)"
#          USEPA_npic$Classification[
 #           str_detect(USEPA_npic$Chemical,
  #                     "Imazalil sulfate")
   #       ],
    
    ,
    TRUE ~ Classification
    )
  ) |> 
  # previous might be deleted later
  
  mutate(NPIC_Carcinogenicity =
           case_when(
             str_detect(Classification, regex("^Carcinogenic Potential|Group B|Group C|^Likely To Be Carcinogenic To Humans|Probably Carcinogenic|Suggestive Evidence|Parent of Imazalil sulfate", ignore_case = TRUE)) ~ "Carcinogenic",
             TRUE ~ NA))|>
   filter(!is.na(NPIC_Carcinogenicity))

  
 

HBRU_npic_carcichems <- left_join(HBRU, USEPA_npic_carcichems|> select(!Chemical) |> 
                                    mutate(CAS_RN = case_when(
                                             CAS_RN == "106325-08-0; 133855-98-8" ~ "133855-98-8",
                                           TRUE ~ CAS_RN))
                                  , by = "CAS_RN") |> 
  
  # next might be deleted later
  mutate(Classification = case_when(
    is.na(Classification) &
        str_detect(Chemical, regex("DMST", ignore_case = TRUE)) ~
          USEPA_npic$Classification[
            str_detect(USEPA_npic$Chemical,
                       "Tolyfluanid")
          ],
    TRUE ~ Classification
    )
  ) |> 
  # previous might be deleted later
  select(1:2,7,10) |> 
  filter(!is.na(Classification))

print(paste("There are", nrow(USEPA_npic), "chemicals listed by NPIC including", nrow(USEPA_npic_carcichems), "with evidence of carcinogenicity and", nrow(HBRU_npic_carcichems), "of them are targeted in HBRU method"))
```

"There are 692 chemicals listed by NPIC including 200 with evidence of carcinogenicity and 35 of them are targeted in HBRU method"


```{r}
USEPA_chems <- full_join (IRIS_chemicals, USEPA_npic, by = "CAS_RN") #|> 
#  filter(
 #   !(str_detect(`CHEMICAL NAME`, "Chlordane") &
  #         str_detect(`WOE DESCRIPTION`, "B2")))

USEPA_carcichems <- full_join(IRIS_carcichems, USEPA_npic_carcichems, by = "CAS_RN") |> 
  select(7,1:2,4,9) 


HBRU_USEPA_carcichems <- left_join(HBRU, USEPA_carcichems |> select(!Chemical)|> 
                                    mutate(CAS_RN = case_when(
                                             CAS_RN == "106325-08-0; 133855-98-8" ~ "133855-98-8",
                                           TRUE ~ CAS_RN))
                                   , by = "CAS_RN") |>
  mutate(
    `WOE DESCRIPTION` = case_when(


      is.na(`WOE DESCRIPTION`) &
        str_detect(Chemical, regex("cis-chlordane|trans-chlordane", ignore_case = TRUE)) ~ "Isomer in racemic mixture of Chlordane (Known/likely human carcinogen)"
      ,
      
      is.na(`WOE DESCRIPTION`) &
        str_detect(Chemical, regex("oxy-chlordane", ignore_case = TRUE)) ~ "Metabolite of several carcinogens: Chlordane (Known/likely human carcinogen) and Heptachlor (B2 (Probable human carcinogen - based on sufficient evidence of carcinogenicity in animals))"
      ,
      
      
      is.na(`WOE DESCRIPTION`) &
        str_detect(Chemical, regex("PCB", ignore_case = TRUE)) ~
          IRIS_carcichems$`WOE DESCRIPTION`[
            str_detect(IRIS_carcichems$`CHEMICAL NAME`,
                       "PCBs")
          ],
      is.na(`WOE DESCRIPTION`) &
        str_detect(Chemical, regex("o,p'-DDD", ignore_case = TRUE)) ~
          IRIS_carcichems$`WOE DESCRIPTION`[
            str_detect(IRIS_carcichems$`CHEMICAL NAME`,
                       "(DDD)")
          ],
      is.na(`WOE DESCRIPTION`) &
        str_detect(Chemical, regex("o,p'-DDE", ignore_case = TRUE)) ~
          IRIS_carcichems$`WOE DESCRIPTION`[
            str_detect(IRIS_carcichems$`CHEMICAL NAME`,
                       "(DDE)")
          ],
        is.na(`WOE DESCRIPTION`) &
        str_detect(Chemical, regex("o,p'-DDT", ignore_case = TRUE)) ~
          IRIS_carcichems$`WOE DESCRIPTION`[
            str_detect(IRIS_carcichems$`CHEMICAL NAME`,
                       "(DDT)")
          ],
      
      is.na(`WOE DESCRIPTION`) &
        str_detect(Chemical, regex("Endrin", ignore_case = TRUE)) ~ "Metabolite of several chemicals: Aldrin (B2 (Probable human carcinogen - based on sufficient evidence of carcinogenicity in animals)), Dieldrin (B2 (Probable human carcinogen - based on sufficient evidence of carcinogenicity in animals)) and Isodrin"
      ,
           

      TRUE ~ `WOE DESCRIPTION`
    ),
    
    Classification = case_when(
      Classification == "See: Imazalil Sulfate, (PC Code 111902)." &
        str_detect(Chemical, regex("^Imazalil$", ignore_case = TRUE)) ~
          USEPA_npic$Classification[
            str_detect(USEPA_npic$Chemical,
                       "Imazalil sulfate")
          ],
      is.na(Classification) &
        str_detect(Chemical, regex("DMST", ignore_case = TRUE)) ~ "Metabolite of Tolyfluanid (Likely To Be Carcinogenic To Humans.)",
      
      is.na(Classification) &
        str_detect(Chemical, regex("dichlorophenyl", ignore_case = TRUE)) ~ "Metabolite of Diuron (Likely To Be Carcinogenic To Humans)"
      ,
      
      is.na(Classification) &
        str_detect(Chemical, regex("Fipronil sulfone", ignore_case = TRUE)) ~ "Metabolite of Fipronil (Group C--Possible Human Carcinogen.)"
      ,            
      
      is.na(Classification) &
        str_detect(Chemical, regex("3-PBA", ignore_case = TRUE)) ~ "Metabolite of several chemicals: Permethrin (Suggestive Evidence Of Carcinogenic Potential), Cypermethrin (Group C--Possible Human Carcinogen), Deltamethrin, Cyhalothrin and Fenvalerate"
      ,   
       
      is.na(Classification) &
        str_detect(Chemical, regex("dichloroaniline", ignore_case = TRUE)) ~ "Metabolite of several carcinogens: Diuron (Likely To Be Carcinogenic To Humans) and Linuron (Group C--Possible Human Carcinogen)"
      ,
       
      is.na(Classification) &
        str_detect(Chemical, regex("DMP|DMTP|DMDTP", ignore_case = TRUE)) ~ "Metabolite of several chemicals: Malathion (Suggestive Evidence Of Carcinogenicity, But Not Sufficient To Assess Human Carcinogenic Potential), Dimethoate (Group C--Possible Human Carcinogen) and other dimethyl-OPs"
      ,
      
      is.na(Classification) &
        str_detect(Chemical, regex("DEP|DETP", ignore_case = TRUE)) ~ "Metabolite of several chemicals: Parathion (Group C--Possible Human Carcinogen), Diazinon and Chlorpyrifos"
      ,

      is.na(Classification) &
        str_detect(Chemical, regex("DEDTP", ignore_case = TRUE)) ~ "Metabolite of several chemicals: Malathion (Suggestive Evidence Of Carcinogenicity, But Not Sufficient To Assess Human Carcinogenic Potential), Diazinon (2A) and Chlorpyrifos"
      ,
      
      is.na(Classification) &
        str_detect(Chemical, regex("PNP", ignore_case = TRUE)) ~ "Metabolite of several chemicals: Parathion (Group C--Possible Human Carcinogen), Diazinon and Chlorpyrifos"
      ,
            
    TRUE ~ Classification
    )
  ) |> 
  mutate(Chemical = if_else(is.na(Chemical), `CHEMICAL NAME`, Chemical),
         USEPA_Carcinogenicity = if_else(is.na(Classification), `WOE DESCRIPTION`, Classification)
         ) |> 
  select(1:2,7:9) |> 
  filter(!is.na(USEPA_Carcinogenicity))


print(paste("There are", nrow(USEPA_chems), "chemicals listed by USEPA after merging IRIS and NPIC databases, including", nrow(USEPA_carcichems), "chemicals with evidence of carcinogenicity and", nrow(HBRU_USEPA_carcichems), "of them are targeted in HBRU method"))

```

"There are 965 chemicals listed by USEPA after merging IRIS and NPIC databases, including 326 chemicals with evidence of carcinogenicity and 57 of them are targeted in HBRU method"

## 1.2. Cross-referencing carcinogenicity

```{r}

left_join(HBRU |> 
            select(1:2),
          HBRU_IARC_carcichem |> 
            select(!1),
          by = "CAS_RN") |> 
  left_join(HBRU_ECHA_carcichem |> 
            select(!1),
          by = "CAS_RN")  |> 
  left_join(HBRU_USEPA_carcichems |> 
            select(!c(1,3:4)),
          by = "CAS_RN")  |> 
  left_join(HBRU_DFG_carcichem |> 
            select(2,7),
          by = "CAS_RN")  |> 
  left_join(HBRU_HCIS_carcichem |> 
            select(!c(1,3:4)),
          by = "CAS_RN")   |> 
  left_join(PubChem_carcichems |> 
            select(!1),
          by = "CAS_RN") |>
  mutate(
    Cross_referenced_Carcinogenicity = case_when(


      # 1. Metabolite of several chemicals
      if_any(all_of(3:8),
             ~ str_detect(.x, regex("Metabolite of several chemicals",
                                    ignore_case = TRUE))) ~
        "Metabolite of several chemicals (including carcinogens)",
      
      # 2. Metabolite of several carcinogens
      if_any(all_of(3:8),
             ~ str_detect(.x, regex("Metabolite of several carcinogens",
                                    ignore_case = TRUE))) ~
        "Metabolite of several carcinogens",

      
      # 3. Metabolite of a carcinogen
      if_any(all_of(3:8),
             ~ str_detect(.x, regex("Metabolite of",
                                    ignore_case = TRUE))) ~
        "Metabolite of a carcinogen",

      # 4. Parent of a carcinogen
      if_any(all_of(3:8),
             ~ str_detect(.x, regex("Parent of",
                                    ignore_case = TRUE))) ~
        "Parent of a carcinogen",
      
      # 5. Carcinogen in humans
      IARC_Carcinogenicity == "1" |
        str_detect(ECHA_Carcinogenicity, "H350") |
        str_detect(HCIS_Carcinogenicity, "H350") |
        str_detect(PubChem_Carcinogenicity, "H350") |
        DFG_Carcinogenicity %in% c(1, 2) |
        str_detect(
          USEPA_Carcinogenicity,
          "^Carcinogenic to humans|^Carcinogen to humans|^Known|Group A"
        ) ~ "Carcinogen in humans",

      # 6. Any evidence from any source
      if_any(
        c(
          IARC_Carcinogenicity,
          ECHA_Carcinogenicity,
          USEPA_Carcinogenicity,
          DFG_Carcinogenicity,
          HCIS_Carcinogenicity,
          PubChem_Carcinogenicity
        ),
        ~ !is.na(.)
      ) ~ "Carcinogen in other models",

      # 7. No information
      TRUE ~ NA_character_
    )
  ) |> 
  distinct(Chemical, .keep_all = TRUE) -> Crossreferenced_carcinogenicity


Crossreferenced_carcinogenicity |> 
  group_by(Cross_referenced_Carcinogenicity) |> 
  summarize(n = n()) |> 
  gt()

print(paste("HBRU method targets 153 chemicals including", 
            nrow(Crossreferenced_carcinogenicity |> filter(Cross_referenced_Carcinogenicity == "Carcinogen in humans")),
            "carcinogens in humans, ",
            nrow(Crossreferenced_carcinogenicity |> filter(Cross_referenced_Carcinogenicity == "Carcinogen in other models")),
                 "carcinogens in other models, ",
            nrow(Crossreferenced_carcinogenicity |> filter(Cross_referenced_Carcinogenicity == "Parent of a carcinogen")),
                 "parent specific of a unique carcinogen, ",
            nrow(Crossreferenced_carcinogenicity |> filter(Cross_referenced_Carcinogenicity == "Metabolite of a carcinogen")),
                 "metabolites specific of a unique carcinogen, ",
            nrow(Crossreferenced_carcinogenicity |> filter(Cross_referenced_Carcinogenicity == "Metabolite of several carcinogens")),
                 "metabolites of several carcinogens and",
            nrow(Crossreferenced_carcinogenicity |> filter(Cross_referenced_Carcinogenicity == "Metabolite of several chemicals (including carcinogens)")),
                 "metabolites of several chemicals (including carcinogens) with the rest",
            nrow(Crossreferenced_carcinogenicity |> filter(is.na(Cross_referenced_Carcinogenicity))),
                 "chemicals unclassified."
                 ))
```

# 2. Combining information of carcinogenic chemicals with their cancer site

## 2.1. Importing list of carcinogenic chemicals with cancer site
```{r}
chems_cancer_site <- read_excel("data/carcicems_cancer-site.xlsx") |> 
  rename(CAS_RN = "CAS n°",
         Chemical_cancersite = "Chemical",
         Cancer_site = "Cancer") |> 
  select(1:2,10:11)
```

## 2.2. Merging the cross referenced list with the cancer site list and assignig cancer site to specific metabolites or metabolites to several carcinogens with same cancer site (oxy-chlordane)
```{r}
carcinogenicity_cancersite_relatedchem <- left_join(Crossreferenced_carcinogenicity,
          chems_cancer_site |> 
            select(1,3),
          by = "CAS_RN") |> 
  left_join(HBRU_relatedchems |> 
              select(2,4),
            by = "CAS_RN")  |>
  filter(!duplicated(Chemical))

parentchem_cancersite <- carcinogenicity_cancersite_relatedchem |> 
  select(Chemical, Parent_Cancer_site = Cancer_site)

Crossreferenced_carcinogenicity_cancersite_metabolites <- left_join(
  carcinogenicity_cancersite_relatedchem,
  parentchem_cancersite,
  by = c("Related chemical" = "Chemical")
  )  |> 
  mutate(
    Cancer_site = case_when(
      Chemical == "oxy-chlordane" ~
        "Liver tumour in mice (same for Chlordane and Heptachlor)",

      Cross_referenced_Carcinogenicity == "Parent of a carcinogen" &
        is.na(Cancer_site) ~
        Parent_Cancer_site,

      Cross_referenced_Carcinogenicity == "Metabolite of a carcinogen" &
        is.na(Cancer_site) ~
        Parent_Cancer_site,

      TRUE ~ Cancer_site
      )
    )|> 
  select(!Parent_Cancer_site)

Crossreferenced_carcinogenicity_cancersite_metabolites |> 
  gt()
 
```


```{r}
write.xlsx(Crossreferenced_carcinogenicity_cancersite_metabolites |> arrange(Chemical), "exported/Crossreferenced_carcinogenicity_cancersite_metabolites.xlsx", rowNames = TRUE)
```

```{r}
Crossreferenced_carcinogenicity_cancersite_metabolites |> 
  filter(!is.na(Cross_referenced_Carcinogenicity))
  
```

