---
title: "Proposed approach for the Characterization"
author: "Franco Catuogno"
format: html
editor: visual
---



```{r setup}
#| include: false
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

# Required packages
```{r}
library(readxl)
library(dplyr)
library(readr)
library(stringr)
library(pdftools)
library(tabulapdf)
library(httr2)
library(tidyverse)
```

# 1. Consulting reference agencies/data bases to identify carcinogens among the targeted chemicals.

## 1.1. Loading lists of carcinogenic chemicals from each agency/data base

### IARC. Downloaded from: https://monographs.iarc.who.int/list-of-classifications/.
```{r}
IARC_chemicals <- read_xlsx("data/Agents Classified by the IARC Monographs, Volumes 1–142.xlsx",
          skip = 1) |> 
  rename(CAS_RN = `CAS No.`) |> 
  filter(!is.na(CAS_RN))

print(paste("There are", nrow(IARC_chemicals), "chemical agents listed by IARC. These number might include repetitions of the same chemical listed with different Agent name"))

```

### ECHA. Downloaded from: https://chem.echa.europa.eu/obligation-lists/clhList.
```{r}
ECHA_H350andH351 <- read_excel("data/Harmonised_List_2026-07-21 12_20_52.xlsx") |> 
  rename(CAS_RN = "CAS number") |> 
  filter(`Hazard class, category and statement code(s)` |> 
           str_detect("H350|351"))

print(paste("There are", nrow(ECHA_H350andH351), "chemicals with Carcinogenicity properties listed by ECHA"))

```
### USA.

```{r}
download.file(
  "https://npic.orst.edu/chemicals_evaluated.pdf",
  destfile = "data/chemicals_evaluated.pdf",
  mode = "wb"
)

library(pdftools)

pdf <- pdf_text("data/chemicals_evaluated.pdf")


length(pdf)

pdf[[13]]
cat(pdf[[13]])

```


```{r}
parse_epa_page <- function(page_number){
  page <- pdf_data("data/chemicals_evaluated.pdf")[[page_number]] |>
    filter(y > 100, !str_detect(text, "Page"))
  
  # --- Step 0: drop lone header-artifact tokens (case-sensitive, exact match) ---
  # These are the wrapped remainders of "PC CODE" / "REPORT DATE" that can land
  # on the same y-coordinate as the first real data row.
  page <- page |>
    filter(!text %in% c("CODE", "DATE"))
  
  # --- Step 1: remove remaining whole-line header/banner rows ---
  header_pattern <- paste(
    "U\\.S\\. Environmental",
    "Protection Agency",
    "CANCER CLASSIFICATION",
    "^CHEMICAL$",
    sep = "|"
  )
  
  line_text <- page |>
    arrange(y, x) |>
    group_by(y) |>
    summarise(line = paste(text, collapse = " "), .groups = "drop")
  
  header_ys <- line_text |>
    filter(str_detect(line, header_pattern)) |>
    pull(y)
  
  page <- page |> filter(!y %in% header_ys)
  
  # --- Step 2: column split + pivot ---
  df <- page |>
    mutate(
      column = case_when(
        x < 160 ~ "Chemical",
        x >= 160 & x < 230 ~ "CAS_RN",
        x >= 230 & x < 270 ~ "PC_Code",
        x >= 270 & x < 530 ~ "Classification",
        x >= 530 ~ "Report_Date"
      )
    ) |>
    arrange(y, x) |>
    group_by(y, column) |>
    summarise(text = paste(text, collapse = " "), .groups = "drop") |>
    pivot_wider(names_from = column, values_from = text) |>
    arrange(y) |>
    mutate(row_id = row_number())
  
  # --- Step 3: collapse remaining wrapped-line fragments into their record ---
  smart_paste <- function(x) {
    x <- x[!is.na(x)]
    if (length(x) == 0) return(NA_character_)
    out <- x[1]
    for (i in seq_along(x)[-1]) {
      out <- if (str_detect(out, "-$")) paste0(out, x[i]) else paste(out, x[i])
    }
    out
  }
  
  df <- df |>
    mutate(
      has_cas = !is.na(CAS_RN),
      grp     = rev(cumsum(rev(has_cas)))
    )
  
  df |>
    group_by(grp) |>
    summarise(
      Chemical       = smart_paste(Chemical),
      CAS_RN         = first(na.omit(CAS_RN)),
      PC_Code        = first(na.omit(PC_Code)),
      Classification = smart_paste(Classification),
      Report_Date    = first(na.omit(Report_Date)),
      first_row      = min(row_id),
      .groups = "drop"
    ) |>
    arrange(first_row) |>
    select(-grp, -first_row)
}

epa13 <- parse_epa_page(13)
```


# Probablemente cuando haga los joints con mi lista the quimicos de HBRU me convenga ponerlos a la izquierda y hacer left join, y ver si se mantienen los replicados de la lista de la derecha
```{r}

```

